CLOCK evolved in cnidaria to synchronize internal rhythms with circadian environmental cues

Version 1

Abstract

Abstract

The circadian clock enables anticipation of the day/night cycle in animals ranging from cnidarians to mammals. Circadian rhythms are generated through a transcription-translation feedback loop (TTFL) involving negative and positive factors. CLOCK is a central conserved positive factor in the animal kingdom. However, the functional evolutionary origin and mechanism of action of CLOCK in basal animals are unknown. In cnidarians, the transcription of putative core clock genes, including Clock, is arrhythmic under constant conditions, and it is unclear how the TTFL model can regulate physiological rhythms. Here, we used the CRISPR/Cas9 system to generate a Clock mutant (Clock-/-) in Nematostella vectensis. High-throughput video tracking of locomotor activity in Clock-/- adults revealed that light regulates rhythmic behavior, while CLOCK maintains a 24-hour circadian rhythm under constant light conditions, and even under a 6-hour cycle, reminiscent of the tidal rhythm. Transcriptome profiling identified hundreds of differentially expressed genes in Clock-/- adults, particularly genes involved in cell cycle and neural development. In accordance with the rhythmic behavioral profile, computational analysis revealed rhythmic and arrhythmic gene expression in wild-type and Clock-/- adults, respectively, under constant dark conditions. Notably, hierarchal light-and Clock-signaling is required to drive rhythmic transcription of the core clock genes. These findings demonstrate that CLOCK has evolved in cnidarians to maintain synchronized 24-hour rhythmic physiology and behavior despite the chaotic inputs of environmental cues.

Introduction

Throughout the history of life on Earth, organisms have had to adapt to a constantly changing environment, including the ∼24-hour daily rhythm of light/dark, driving the development of endogenous biological clocks. The circadian clock, which is entrained by external stimuli such as light, enables the organism to anticipate the onset of the light and dark phases and synchronize its physiology and behavior in harmony with the environment. This, in turn, enhances the organism’s fitness and survival1–3. From single-celled organisms to metazoans, circadian clocks have evolved multiple times, highlighting their importance to living organisms1,2. Despite the fundamental role circadian clocks play in regulating the biorhythmicity of living organisms, their evolutionary origin and intricate molecular mechanisms remain ambiguous in early diverging animal lineages, such as cnidaria.

Rhythmic phenomena, including calcification, reproduction, and diel behavior patterns, have been examined in cnidarian species3–8. While some of these patterns were found to be directly triggered by environmental stimuli, such as light, others persist in the absence of external cues, suggesting the presence of an internally generated and self-sustaining circadian clock6,9,10. At the molecular level, cnidarians possess homologs of putative core clock genes found in bilaterians11–13. Several studies have shown that most of these genes display diel expression patterns under light/dark cycles. However, in contrast to most animals, their oscillation generally ceases without the presence of light cues14–16. Thus, how the core clock genes orchestrate rhythmic gene expression and circadian behaviors in cnidarians remains unclear.

One of the most studied cnidarian species in the field of chronobiology is the estuarine sea anemone, Nematostella vectensis. Few studies have shown that in diel lighting, the locomotor behavior of N. vectensis has a ∼24-h rhythm that is maintained under constant conditions suggesting it is regulated by an endogenous circadian clock6,17–19. In accordance, the N. vectensis genome codes for conserved core clock genes such as Clock, Cycle, Timeout, and the cryptochromes Cry1a and Cry1b12,13. The proposed circadian clock model in N. vectensis is composed of the positive transcription factors, CLOCK and CYCLE, that heterodimerize and upregulate light-dependent cryptochrome genes in the feedback loop, and PAR-bZIPs in the feed-forward loop, which repress the transcription of the positive elements12. More recently, the clock-interacting pacemaker, CIPC was predicted to act as an additional repressor of the CLOCK:CYCLE dimer19. However, in contrast to the free-running oscillation demonstrated for N. vectensis behavior6,17,19, transcriptional expression profiles of most candidate genes implicated in the core clock mechanism do not retain their oscillation period in the absence of light14,15,19.

Here, we employed the CRISPR/Cas9-mediated genome editing system to establish a Clock mutant (Clock-/-) N. vectensis. Using behavioral monitoring and transcriptomic analysis, we studied the role of CLOCK in regulating rhythmic locomotor activity and gene expression under various light regimes. The findings show evolutionary conserved roles for CLOCK and suggest that Clock has evolved to modulates the ancient hierarchy between light-and circadian clock-regulated physiology and behavior.

Results

Phylogenetic analysis and spatial expression of Clock mRNA in N. vectensis

Phylogenetic analysis of CLOCK protein sequences showed that it is positioned within the cnidarian branch (Fig.1a). Similar to the protein structure found in other animals, it contains a basic helix–loop–helix (bHLH) DNA binding domain and two Per-Arnt-Sim (PAS) domains20,21. PAS domains have been shown as important structural motifs involved in protein-protein interactions that drive the self-sustaining molecular mechanism underlying the circadian clock. To localize Clock expression in the cellular and tissue levels, in situ hybridization chain reaction (HCRv.3) was performed using probe designed to target the Clock mRNA. In WT juvenile four-tentacle stage, low clock expression was observed throughout the animal tissue, and enriched expression was visible in the tentacle endodermis and mesenteries (Fig.1b). This expression pattern resembled the expression observed at late planula stage22. Notably, the HCR technique enables to detect low number of transcripts at cellular resolution. Although the expression of Clock was low, Clock transcripts localized in discrete subcellular regions in the nuclei and soma (Fig.1b). These results characterize the spatial expression of Clock with a cellular resolution in the polyp, however, to date, functional manipulation of the Clock gene has not been performed in early animal lineages including cnidarians, and its function is unknown in cnidarians20,21.

Clock is necessary to maintain circadian behavior.

(A) Phylogenetic tree showing the evolutionary relationship of CLOCK orthologs across different animal species. (B) In situ hybridization of Clock in the WT juvenile with scale bars representing 0.1mm and 5μm. (C) Schematic representation of the Clock gene in grey, with the open reading frame (ORF) in orange and the conserved protein domains bHLH (green) and PAS1 and PAS2 (dark red). The CRISPR generated allele has a 20nt insertion after the PAS1 domain, represented by a black arrowhead. (D-G) Normalized locomotor activity (a.u), hourly binned over 72h, under different light conditions: 12h light:12h dark, continuous dark, continuous light, and 6h light:6h dark. The black line represents the WT and the red line represents the Cloc/- mutant. (D’-G’) Periodograms for each corresponding light condition. The significant period value (p<0.01) is indicated for each genotype in the top left corner of each graph.

Generation of Clock-/- N. vectensis

To investigate the function of CLOCK, we utilized the CRISPR-Cas9 system to generate a stable Clock mutant N. vectensis. The CRISPR guide RNA (gRNA), designed to target a region close to the 5′-end of the Clock coding sequence (CDS), was synthesized, and microinjected with the endonuclease Cas9 into zygotes (Methods). F0 animals were outcrossed with WT and the F1 progeny were raised to adulthood. Subsequently, F1 animals harboring mutation at the Clock locus were identified by genotyping using DNA extracted from bisected physa. We identified 10 different alleles, 6 with a frame-shift mutation including one of 20 bp (Clock1), resulting in a premature stop codon (Extended Data Fig. 2). The predicted 203 amino acid truncated protein lacked 459 amino acids including one co-factor dimerization PAS domains (Fig. 1c). To obtain Clock-/- homozygous animals, we crossed Clock1+/- F1 animals. Genotyping F2 animals resulted in the expected 25% Clock-/- animals. Next, we intercrossed Clock-/- animals to obtain F3 Clock-/- N. vectensis that were used in subsequent experiments, aiming to assess the role of CLOCK in regulating diel behavioral and genetic rhythms.

Clock is necessary to maintain circadian behavior

Our behavior tracking analysis showed that the locomotor activity of both WT and Clock-/- N. vectensis has a periodicity of 24h (p<0.01) under 12h light: 12h dark (LD) cycles (Fig.1d-d’, Supplementary Table 1). Next, to determine whether the Clock mutation influences free-running locomotor oscillations, animals were monitored under constant conditions i.e., continuous dark (DD) or continuous light (LL). We discovered that WT animals maintained the rhythmic behavior with a 22h (p<0.01) periodicity. In contrast, the locomotor activity of Clock-/- N. vectensis was arrhythmic (Fig.1e-e’ and Fig.1f-f’, Supplementary Table 1). To further explore the rhythmicity of N. vectensis under LD, we tracked the locomotor activity under a 6h light: 6h dark (LD 6:6) cycle after a 72h entrainment under 12:12 LD. We found that the WT behavior maintained a weak but significant periodicity of 22h (p<0.01). Interestingly, the Clock-/- N. vectensis showed a clear rhythmic behavior with a 12h period (p<0.01), pointing towards direct light-dependency (Fig.1g-g’). However, to rule out the effect of a potential endogenous circatidal clock, we entrained the animals under LD 6:6 and tracked their behavior under DD. We found that the WT maintained a free-running periodicity of 22h (p<0.01), while the Clock-/- N. vectensis were arrhythmic (Extended Data Fig.1, Supplementary Table 1).

Clock mutation causes a broad impact on the rhythmic transcriptome

To investigate the underlying molecular correlate of the arrhythmic behavior phenotype found in Clock-/- N. vectensis, transcriptional profiling was conducted on WT and Clock-/- animals that were sampled in 4h intervals over 24h under LD and DD (Fig.2a). Using the RAIN (rhythmicity analysis incorporating non-parametric methods) algorithm23, we identified in the WT 2262 genes rhythmic under LD and 2361 rhythmic genes under DD (p< 0.05). In Clock-/-, we detected 1847 rhythmic genes under LD and 1314 under DD (Fig.2b, Extended Data Fig.2, Supplementary Table 2). Overall, we detected 6200/6803 (91.1%) genes with no overlap between genotypes, 3798 (55.8%) genes specific to the WT, and 2402 (35.3%) genes to Clock-/- N. vectensis, revealing that the Clock mutation has a broad impact on the rhythmic transcriptome (Extended Data Fig.2, Supplementary Table 2). To investigate genes strictly under the control of the circadian clock we crossed the two WT gene lists (LD and DD) and obtained 220 genes rhythmic in both LD and DD (Fig.2b, Supplementary Table 3), we refer to them as clock-controlled genes (CCGs). A GO term analysis showed enrichment in regulation of different step of translation, transcription, and mitosis regulation (Fig.2d, Supplementary Table 4) which is coherent with the known function of the N. vectensis circadian clock14. In Clock-/-, 188/220 (85%) of the CCGs were not rhythmic under LD nor DD. As such, we have designated them as Clock-dependent CCGs (Fig.2c, Extended Data Fig.2, Supplementary Table 3). Next, we analyzed the Clock-dependent CCGs for DNA motif enrichment in their promoter, 1kb upstream of their ATG, and detected a significant enrichment in multiple E-box motif associated with bHLH circadian regulation such as CLOCK, BMAL1, and NPAS2 in 28% of the genes (Fig.2e). Applying the same analytical methods, we discovered that in Clock-/- N. vectensis, 125 out of 156 genes displayed rhythmic expression patterns under both LD and DD. These genes were arrhythmic in the WT animals (Fig.2c, Extended Data Fig.2 and Supplementary Table 3). As such, we have designated these genes as “neoCCGs”. We did not observe any significant enrichment of gene ontology (GO) terms related to circadian rhythms or the presence of circadian E-box motifs in the promoter regions of the neoCCGs. (Fig.2d-e and Supplementary Table 4).

Clock mutation causes a broad impact on the rhythmic transcriptome.

(A) Overview of the experimental design used to generate RNA-seq data. Polyps were entrained for 72 hours before sampling at 4-hour intervals over a 24-hour period (dark arrows) in both LD and DD cycles. (B) Venn diagram comparing the total number of rhythmic genes identified in WT and Clock-/-in LD and DD cycles. (C) Heatmap showing genotype-specific rhythmic transcripts in WT and Clock-/-. (D) Gene Ontology (GO) terms that were significantly enriched (Bonferroni corrected p < 0.05) for rhythmic genes in WT. (E) De novo sequence motif analysis of circadian transcriptional regulators detected in WT and Clock-/- using Homer/MEME. The 1000bp upstream ATG of each rhythmic gene was used for motif analysis.

Clock cooperates with light to drive the transcription rhythm of the core clock genes

Consistent with previous finding in N. vectensis 14,15,19, our rhythmic analysis of the eight core clock candidates showed that they are all arrhythmic under DD (Fig.3a and Supplementary Table 2). This finding was also true for the Clock-/- N. vectensis, however, we observed general altered expression patterns under LD (Fig.3a). Interestingly, PAR-bZIPc and Cry-b were found to be rhythmic only in WT LD, indicating that light is epistatic on Clock (Fig.3a). Notably, the only core clock member included in the Clock-dependent CCGs we identified, was CiPC, however, its individual expression pattern suggested that it is more likely a dampening factor rather than a true rhythmic clock gene (Fig.3a). Additionally, our analysis revealed that circadian E-box motifs were enriched in most of the core clock genes. (Fig.3b). Given the limitations in using core clock genes mRNA as a true circadian clock marker, we sought to identify new markers that could accurately reflect the circadian rhythm. Through our analysis of the 188 Clock-dependent CCGs, we identified 16 genes that displayed rhythmic expression patterns in both our study and a previous study by Leach et al. (2019)14 (Fig.3c, Supplementary Table 5). Among these candidate markers, we selected Myh7 due to its detectability through both RNA-seq and in situ hybridization techniques. Furthermore, the spatial expression pattern of Myh7 closely resembled that of Clock, as evidenced by our analysis (Fig.3d-f).

Clock cooperates with light to drive the transcription rhythm of the core clock genes.

(A) Each core clock gene is individually plotted showing the read count over 24h in LD and DD in WT (black) and Clock-/- (red). The continuous line represents significant rhythmicity (RAIN p<0.05) while the dashed line indicates no rhythmicity. (B) The presence of Circadian E-box in all core clock gene promoters was detected using HOMER within 1000bp upstream ATG. (C) Conserved rhythmic genes were selected from our WT data and Leach et al., 2020. (D) The temporal expression pattern of Clock-downstream gene Myh7 is shown. (E) In situ hybridization of Myh7 in the WT juvenile with scale bars representing 0.1mm. Nuclei are DAPI-stained, colored in grey.

Clock regulates genetic pathways

To further evaluate changes in gene expression between the genotypes caused by the Clock mutation, differentially expressed gene analysis was performed between genotypes for each light condition. Differentially expressed genes were identified using the Wald test with an adjusted p-value cutoff of 0.05. By analyzing the differentially expressed genes from each light condition, and comparing the genes expressed by all Clock-/- samples in comparison with WT samples, we aimed to identify processes regulated by Clock. In LD conditions, Clock-/- N. vectensis showed 457 down-regulated genes and 646 up-regulated genes, without any significant GO term enrichment (Fig.4a, Supplementary Table 5-6). In DD conditions, Clock-/- displayed 2450 down-regulated genes and 1770 up-regulated genes (Fig.4b, Supplementary Table 5). Interestingly, the down-regulated genes showed GO term enrichment in various cellular processes such as spermatid development, mitotic and meiotic cell divisions, and ciliary/flagellar motility. In contrast, we observed that the up-regulated genes were significantly enriched in processes such as the modulation of another organism’s processes, sensory perception, and ion transmembrane transport (Fig.4c, Supplementary Table 6).

Clock regulates genetic pathways.

(A) Volcano plot showing the differential expression of genes between WT and Clock-/- in LD. The dashed lines indicate the threshold used to detect DEGs (Padj<0.05) and the WT and Clock-/- are represented by black and red dots, respectively. (B) Volcano plot showing the differential expression of genes between WT and Clock-/- in DD (C) Gene Ontology (GO) terms with significant fold-enrichment (Bonferroni corrected p-value < 0.05) are shown for DEG analysis in DD.

Discussion

This study offers valuable insights into the evolutionary history of circadian clocks by demonstrating the effects of a loss-of-function Clock mutation in the basal metazoan, Nematostella vectensis. Through our behavioral assays, we have shown that the Clock gene is essential for maintaining rhythmic locomotor activity in the absence of an entraining light cue. These findings suggest that this fundamental aspect of circadian clocks was likely present in the common ancestor of both N. vectensis and bilaterian animals, shedding light on the evolution of this critical feature of animal biology. However, our transcriptomic data raises questions regarding the ancestral role of Clock in the central clock mechanism and its relationship with the environment.

Studies of homozygous Clock-/- flies have shown that dClock has a dual role in generating circadian rhythms: as an essential core component of the molecular oscillator and in influencing the direct effects of light on locomotor activity24,25. WT flies under standard light-dark cycles anticipate the lights-on and lights-off transitions by slowly increasing activity, while this anticipatory response is obscured in arrhythmic dClock mutants26–28. In contrast to flies, CLOCK-deficient mice continue to express robust circadian rhythms in locomotor activity, although they do show altered responses to light29,30. Overall, these findings suggest that CLOCK has different roles in the generation of circadian rhythms and in modulating responses to light in flies and mice. Interestingly, Bmal1 is the only known mammalian clock gene whose inactivation results in immediate arrhythmicity at both the behavioral and molecular levels in the absence of external cues31. However, since mammalian genes exist as multiple isoforms with partially overlapping functions, it was suggested that functional redundancy may obscure the phenotypic outcome in the case of other clock components, including mClock32–34. The results presented here are consistent with these previous findings, as observed in the complete loss of rhythmic circadian behavior in Clock-/- N. vectensis under free-running conditions. However, when exposed to LD cycles, we found that the period of rhythmic behavior in these mutants was light-dependent (Fig.1d-g), indicating that while the molecular oscillator in Clock-/- is disrupted, the light input pathway remains intact. Thus, Clock is essential for maintaining circadian behavior in the absence of an entraining cue; however, unlike dClock24,25,28 or mMop331 (also known as BMAL1), it does not seem to play a functional role in the input pathway as it is not necessary for entraining or generating normal activity levels of behavior to LD cycles.

At the transcriptomic level, previous studies in N. vectensis have shown large shifts in the transcriptional profile of many genes after a single day of constant darkness, including the candidate core circadian genes that were found arrhythmic12,14,22. Similarly, in our data, the rhythmic expression of these genes, including Clock, strongly depends on light-dark cycles while behavior and CCGs are still oscillating in free-running. One explanation for this inconsistency is that using whole animals for sampling material, as in vertebrates35, has the potential to mask oscillating gene expression signals, particularly if the signals are present in a small number of cells or If tissues have rhythmic gene expression in different phases10,14. Another possibility is that the core clock mechanism in cnidarians may differ from that of other animal groups, which could affect the intrinsic transcriptional rhythmicity. However, given the widespread expression pattern of Clock mRNA we detected, we chose to perform full animal sampling for our transcriptome analysis (Fig.1b).

Even though LD cycles are the main factor driving rhythmic expression, all eight core clock genes have an altered expression pattern in our Clock-/- (Fig.3a). This finding could suggest that Clock exerts a fine-tuning regulation on their light-dependent expression. However, while the actual contribution of the transcriptional oscillation of putative core clock genes to the oscillator, if any, remains indefinite, the observation of transcriptomic circadian oscillations and the identification of 125 neoCCGs in Clock-/- raises further unexpected questions. The only study describing a similar phenomenon is in liver cells of Bmal1-/- mice, which exhibit circadian oscillations at the transcription, translation, and protein phosphorylation levels. The authors proposed that a non-canonical clock driven by ETS factors is functioning36. Nevertheless, in our data we did not retrieve any GO term enrichment, suggesting the neoCCGs are not recruited by the oscillator in a coordinated manner. Additionally, we did not identify circadian annotated E-box enrichment within their promoter, nor other binding motif enriched compared to the CCGs which could help identify which factors are responsible for these abnormal oscillations. Although we cannot exclude the possibility of an effect from the predicted truncated CLOCK protein, the rhythmic circadian behavior of the heterozygote Clock-/+ indicates there is no dominant-negative effect (Extended Data Fig.1). From our results, it is apparent that a functional circadian oscillator persists in the absence of Clock, albeit with altered target genes. It appears that Clock is not essential for the core mechanism that drives molecular oscillations. However, the origin of these oscillations remains unclear, and further research is necessary to elucidate the evolutionary and mechanistic implications of our findings.

Our study revealed that the Clock-/- gene exhibits significant differential expression when compared to the WT in LD and DD conditions. Interestingly in DD, the down-regulated genes were mainly involved in cell division and microtubule organization, which suggests a potential function of Clock in coordinating cellular processes beyond circadian rhythm. Conversely, the up-regulated genes were linked to sensory perception, indicating a potential novel role of Clock in regulating sensory processes. Overall, our findings suggest that Clock has non-circadian functions in a light-dependent context in N. vectensis, possibly related to developmental processes, as core clock genes are expressed and arrhythmic during planula stages22. Our study underscores the need to explore potential novel roles of circadian clock components beyond their established role in regulating the circadian rhythm of adult N. vectensis.

Taken as a whole, this study provides novel insights into circadian regulation in Nematostella vectensis and sheds light on the evolutionary origin of circadian time maintenance. Our findings indicate that while Clock function is conserved from Cnidaria to mammals, it is not a core component of the TTFL. Our findings suggest that the CLOCK protein evolved to alter the light-biological clock hierarchy and facilitate the connection between the core oscillator and the organism, simplifying interactions with the environment. This opens new avenues for investigating cell-type-specific mechanisms of the circadian clock that drive the molecular and phenotypical oscillations of cnidarians. By further exploring the circadian clock mechanisms in cnidarians, we can gain deeper insights into the evolutionary origins of this critical aspect of biology, enhancing our understanding of how organisms have evolved to keep track of time and adapt to their environment.

Methods

Animal husbandry

N. vectensis were grown in 12g.L sea salt water at 17°C, and fed with Artemia salina nauplii three times a week. Spawning of gametes and fertilization was performed according to a published protocol37. In brief, temperature was raised to 25 °C for 9 h and the animals were exposed to strong white light. Three hours after the induction, oocytes were mixed with sperm to allow fertilization.

CRISPR/Cas9 mediated mutagenesis

Genome editing in N. vectensis was carried out following establishe CRISPR/Cas9 protocols, with slight modifications38,39. ZiFiT targeting software (http://zifit.partners.org/)40 was used to select a guide RNA (gRNA) target site within the beginning of the Clock exon and to design complementary oligonucleotides. To ensure specificity of the gRNA, the selected target site sequence (GGTCCTCTCGTGGACTCTAC) was BLASTed against Nematostella vectensis genome (using JGI expected E value threshold of 0.1 to adjust for short sequences; http://genome.jgi.doe.gov/Nemve1/Nemve1.home.html). To generate the gRNA template, the following oligonucleotides were used:

Oligo 1: 5’- TAGGTCCTCTCGTGGACTCTAC -3’

Oligo 2: 5’- AAACGTAGAGTCCACGAGAGGA -3’

To construct the gRNA expression vector, pDR274 (plasmid # 42250; Addgene) was digested with the BsaI restriction enzyme. Subsequently, the gRNA oligonucleotides were annealed and cloned into the BsaI-digested pDR274 vector. Next, DraI-digested gRNA expression vectors, purified via ethanol precipitation followed by PureLink PCR purification kit (Invitrogen), were transcribed and purified using HiScribe T7 High Yield Transcription Kit (New England BioLabs) and illustra Microspin G-50 Columns (GE Healthcare Life Sciences), respectively. Cas9 recombinant protein with nuclear localization signal (260 ng/μl; PNA Bio, USA) was co-injected with the gRNA (140ng/µl) into N. vectensis zygotes. Injected embryos were raised in petri dishes at 22°C under constant darkness with daily water changes.

CRISPR/Cas9 mediated mutagenesis screening

To evaluate genome editing efficiency and mosaicism in F0 animals, genomic DNA flanking the guide sequence was amplified and Sanger sequenced. PCR was performed using two strategies. For the first, PCR reactions were performed using individual injected N. vectensis (7-days post-fertilization), directly pipetted into a 25-μl PCR reaction containing a 2x concentration of PCR MasterMix (Tiangen) and 10 pmol of each PCR primer. For the second, genomic DNA was extracted from the physa bisected from live adult animals, after relaxation in 7% MgCl2 (Sigma-Aldrich), using NucleoSpin Tissue DNA purification kit (MACHERY-NAGEL). Subsequent PCR reactions were performed as above using 50-ng of genomic DNA. The primers used for these reactions (listed below) were designed to amplify a ∼750bp region around the targeted Clock genomic locus. Mosaicism was determined if sequenced PCR products showed overlapping peaks in their chromatograms. The second strategy, which takes advantage of the ability of N. vectensis to fully regenerate within a few days41,42, is the one we refer to in the text hereafter. The injected individuals determined mosaic mutants were raised as F0 founders to sexual maturity and outcrossed with wild-type animals. The progeny of these crosses was raised and individually genotyped as described above. To determine inheritable mutations, sequences were further analyzed using the Tracking of Indels by DEcomposition web-tool (TIDE). TIDE quantifies editing efficiency and identifies the predominant types of DNA insertions and deletions (indel) mutation composition from a heterogeneous PCR product compared to a wild-type sequence43. Different heterozygous mutants were raised to sexual maturity and outcrossed with wild-type animals. The resulting F2 progenies were then raised to sexual maturity and genotyped before spawning for F3. Heterozygous mutants from each F2 progeny were intercrossed to obtain 25% homozygous F3 mutants. All animals used in this study are derived from heterozygous F2 mutants intercrosses, harboring the mutant allele Clock1.

PCR genotyping was performed using the following primer:

Forward 5’- GATAAACACGGGCCGAAGATA -3’

Reverse 5’- CAGTCCACGCTGGTCTAAAT -3’

Determination of Clock1 F3 mutant genotypes

Genomic DNA was extracted as described above and used for following PCR and electrophoresis-based genotyping. PCR primers (listed below) encompassing the Clock targeted site were used to produce PCR products of approximately 100 bp. The PCR products were then loaded and migrated by electrophoresis on a 3% Tris-borate-EDTA (TBE) agarose gel supplemented with GelStar Nucleic Acid Gel Stain (Lonza) for approximately one hour. The genotype of each F3 animal was determined by visualizing differences in migration speed of the PCR products caused by the CRISPR/Cas9 genome editing. The homozygous mutant animal (Clock−/−) produces only the larger ∼120 bp amplicon while the wild-type animal (Clock+/+) produces only the lower ∼100 bp amplicon. Animals heterozygous for the deletion (Clock+/−) produce both the larger mutant and the smaller wild-type amplicons. PCR genotyping was confirmed by subsequent DNA sequencing of selected F3 animals.

PCR was performed using the following primer:

Forward 5’- ACCCCACTGAGTGACCTCTT -3’

Reverse 5’- ATACGCCTGCGCTATACACC -3’

Behavioral assays

Locomotor activity of individual N. vectensis were monitored using a lab-made setup equipped with an IP IR camera (Dahua) and a white neon illumination and LD cycles. The camera output 1h mp4 movie files which were AVI converted then stitched. The data collection and analysis were carried out by EthoVision XT8 video tracking software (Noldus information technology, Wageningen, Netherlands). Animals were isolated in wells of six-well plates, each of which was manually defined as a tracking ‘arena’ in the EthoVision software. Center-point detection with gray scaling (detection range of 25–77, contour erosion of 1-pixel, high pixel smoothing) was used to monitor movements, which were calculated according to the change in position of the average center pixel each second.

Illumination was provided by a white neon light with an intensity of 17umol. m.2s1 (+/−2) and did not significantly affect the experimental temperature (20 °C). The illumination cycles were 12: 12 h LD or 6: 6h LD. Parameters were optimized to ensure that organisms were detected throughout the entire observation period.

Behavior analysis

The total distance moved was summed in hourly bins and individually normalized min/max by the software GraphPad Prism 9.4. The average and standard errors were calculated for all tested animals based on the normalized values of each hour. The oscillation frequencies were evaluated based on the average values of each experiment using Fourier analysis-based software LSP with a p<0.01.

RNA-seq experimental design

All polyps were isolated in wells of six-well plates. Then, they were subjected to the 12: 12h LD cycle with 17umol. m.2s1 (+/−2) light intensity during 72 hours for entrainment in an incubator with a stable temperature at 18°C. Subsequently, the polyps were divided into 2 experimental subgroups: 12: 12h LD and Continuous Dark or DD. Sampling began at 7 am (ZT0) and was performed at 4-hour intervals over 24 hours. At each time point, three or four individual polyps were sampled from each experimental group, immediately snap-frozen in liquid nitrogen, and transferred to − 80°C for storage.

RNA extraction, library preparation, and sequencing

Total RNA was extracted from all sampled polyps (n = 96) using TRIzol reagent (Invitrogen). Purified RNA samples were analyzed using a NanoDrop 1000 spectrophotometer (Thermo Fisher Scientific) to assess RNA quantity and 2200 TapeStation (Agilent) to assess RNA quality (RNA integrity number, >8.5). From each of the 96 samples (4 biological replicates in LD and 3 in DD, with the highest-quality extracts × 4 experimental subgroups × 7 time points, 1.5 μg of RNA was sent for library preparation (INCPM mRNAseq protocol) and sequencing at the Weizmann Institute Sequencing Unit, Israel. The libraries were sequenced using the bulk MARS-seq protocol (Jaitin et al. Science, 2014, Keren-Shaul et al., Nature Protocols,2019) on an Illumina NovaSeq 6000, resulting in an average of 17 million single-end reads of 113 bases per sample.

Bioinformatic analysis

First, the unique molecular identifier (UMI) sequence of each read was extracted and placed within the read1 header file using UMI-tools extract (umi_tools v1.1). Next, the reads were mapped onto the N. vectensis genome (NCBI genome GCA_000209225.1) using STAR (v2.6.0a) (Dobin et al., 2013) with default parameters. Mapped reads were then deduplicated based on UMIs using the umi_tools dedup. The mapped reads were sorted by SAMtools (version 1.9). The number of reads per gene were quantified using HTSeq-Count (v0.12.4) (Anders et al., 2015).

Rhythmicity analysis

Rhythmicity in transcript expression was identified using the RAIN package for R23. The RAIN algorithm was run separately for each N. vectensis genotype in each light condition (LD and DD) as one dataset. For each dataset, all replicat es (n=3) for each time point were analyzed as regular time series to detect dai ly oscillating transcripts. We only looked for transcripts with a precise 24-hour period and not a range (e.g., 10 to 14 or 20 to 28 hours). To account for multip le testing, transcript with a Benjamini-Hochberg-corrected RAIN P value < 0.05 were considered confident cyclers. Heatmaps were generated using the heatm ap package (v4.5.5) in R. Venn diagrams were generated using the web tool V enn diagram (http://bioinformatics.psb.ugent.be/webtools/Venn/) And redraw w ith Inkscape. Expression plots were generated using GraphPad Prism (V.9.1).

GO term enrichment analysis

After obtaining the differential gene expression results, Gene Ontology (GO) analysis was performed using the R TopGO package (v2.50.0) to identify the significantly enriched biological processes, cellular components, and molecular functions in the differentially expressed genes. The algorithm assigns a significance score to each GO term based on the enrichment p-value and the specificity of the term. In this study, the GO analysis was performed separately for the up-regulated and down-regulated genes in each condition (LD and DD) to identify the specific biological processes and molecular functions that are affected by the Clock mutation.

DNA motif enrichment analysis

DNA motif promoter enrichment analysis was performed using the Homer software (v4.10) and MEME Suite (v5.1.1). First, sequences for promoter regions (1000kb upstream ATG) of differentially expressed genes were extracted. The resulting fasta files were then used as input to the Homer software to identify enriched DNA motifs. Finally, the identified motifs were also analyzed using the MEME Suite to confirm their statistical significance and obtain their corresponding sequence logos. We manually identified in the list of motif enrichment all the E-box motifs and Circadian E-box motifs. Boxplot were generated using GraphPad Prism version 9.5.1.

Differential Expression Analysis

Differential expression analysis was performed using R (v4.2.2) Bioconductor package, DESeq2 (v1.38.3)44. Raw read counts were obtained using I used HTSeq-Count (v0.12.4)45 and then imported into DESeq2 for normalization and statistical analysis. Differentially expressed genes were identified using the Wald test with an adjusted p-value cutoff of 0.05. The analysis was performed on all the time-points pooled of each genotype per light condition. The output of the analysis includes a list of genes with their log2 fold change, p-value, and adjusted p-value. Volcano-plots were generated using GraphPad Prism version 9.5.1.

HCR v.3 in situ hybridization

A custom Clock (NVE2080, amplifier: B3 and B5) and Myh7 probe set (NVE14552, amplifier: B5) were generated. We used zfHcrt probe set (ZDB-GENE-040324-1, amplifier: B1 and B3) as a negative control. For HCR on Nematostella juvenile, several alterations were made to a previously described protocol46. Briefly, polyps were plucked and fixed in 4% PFA overnight at 4°C. Polyps were washed 3× in 1× PBS and then dehydrated and permeabilized with 2×5 min washes in 100% methanol. The samples were stored at –20°C overnight. To rehydrate the samples, a series of graded MeOH/PBST washes were used for 5 min each: 75% MeOH:25% 1× PBST, 50% MeOH:50% 1× PBST, 25% MeOH:75% 1× PBST, and finally 2× washes in 100% 1× PBST. To further permeabilize the polyps, samples were incubated in 10 ug/ml Proteinase K diluted in 1× PBST for 10 min. Samples were quickly washed 3× in 1× PBST, and then post-fixed with 4% PFA for 10 min. After post-fixation, samples underwent 3×5 min washes with 1× 2xSSC+0.1% Triton. From now, the following solutions (Pre-hybridization, hybridization and probe wash buffers) were lab-made from the cnidarian-adapted hybridization buffer47. Samples were then pre-hybridized with pre-hybridization buffer at 37°C for 30 min. After pre-hybridization, samples were incubated with 2 pmol of the probe set diluted in hybridization buffer for 16 hr at 37°C. To remove the probe mixture solution, samples were washed 2× for 30 min each with probe wash buffer at 37°C. Samples were washed 2× for 5 min with 5× SSC +0.1% Triton and then treated with probe amplification buffer for 30 min at room temperature. Samples were washed into hairpin amplification buffer containing snap cooled amplifier hairpins and were incubated at room temperature, protected from light, overnight. Samples were then washed with successive 3× 5xSSC +0.1% Triton washes: 2× washes for 15min. Nuclear staining was performed using DAPI 1:1000 in PBST during 1h. Samples were then washed with successive 2× 5xSSC +0.1% Triton washes: 2× washes for 5min. Eventually were slide-mounted in glycerol and stored at 4°C.

Microscopy and Image processing

Samples were imaged using a Zeiss LSM 710 with a x63 oil objective. They were slide-mounted in glycerol. Image manipulation was performed with Fiji48. For the double probes Clock imaging (Fig1.B), ROIs were generated from each Clock probes and only the ROIs positive for the 2 fluorophores were kept. These ROIs were then used to extract from the original picture to select the signal considered as true mRNA signal. Figures were then assembled in Inkscape (http://www.inkscape.org/)

Data availability

The RNA-seq data reported in this study have been deposited to the Sequence Read Archive (SRA), under accession PRJNA935092. All data supporting the findings of this study are included in the manuscript and its supplementary files or available upon request from the corresponding authors.

Supporting information

eLife Assessment

The study provides potentially fundamental insight into the function and evolution of daily rhythms. The authors investigate the function of the putative core circadian clock gene Clock in the cnidarian Nematostella vectensis. While it parts still incomplete, the evidence suggests that, in contrast to mice and fruit flies, Clock in this species is necessary for daily rhythms under constant conditions, but not under a rhythmic light/dark cycle, suggesting that the major role of the circadian oscillator in this species could be a stabilizing function under non-rhythmic environmental conditions.

Reviewer #1 (Public Review):

In this nice study, the authors set out to investigate the role of the canonical circadian gene Clock in the rhythmic biology of the basal metazoan Nematostella vectensis, a sea anemone, which might illuminate the evolution of the Clock gene functionality. To achieve their aims the team generated a Clock knockout mutant line (Clock-/- ) by CRISPR/Cas9 gene deletion and subsequent crossing. They then compared wild-type (WT) with Clock-/- animals for locomotor activity and transcriptomic changes over time in constant darkness (DD) and under light/dark cycles to establish these phenotypes under circadian control and those driven by light cycles. In addition, they used Hybridization Chain Reaction-In situ Hybridization (HCR-ISH) to demonstrate the spatial expression of Clock and a putative circadian clocl-controlled gene Myh7 in whole-mounted juvenile anemones.

The authors demonstrate that under LD both WT and Clock-/- animals were behaviourally rhythmic but under DD the mutants lost this rhythmicity, indicating that Clock is necessary for endogenous rhythms in activity. With altered LD regimes (LD6:6) they show also that Clock is light-dependent. RNAseq comparisons of rhythmic gene expression in WT and Clock-/- animals suggest that clock KO has a profound effect on the rhythmic genome, with very little overlap in rhythmic transcripts between the two phenotypes; of the rhythmic genes in both LD and DD in WT animals (220- termed clock-controlled genes, CCGS) 85% were not rhythmic in Clock-/- animals in either light condition. In silico gene ontology (GO) analysis of CCGS reflected process associated with circadian control. Correspondingly, those genes rhythmic in KO animals under DD (here termed neoCCGs) were not rhythmic in WT, lacked upstream E-box motifs associated with circadian regulation, and did not display any GO enrichment terms. 'Core' circadian genes (as identified in previous literature) in WT and Clock-/- animals were only rhythmic under entrainment (LD) conditions whilst Clock-/- displayed altered expression profiles under LD compared to WT. Comparing CCGs with previous studies of cycling genes in Nematostellar, the authors selected a gene from 16 rhythmic transcripts. One of these, Myh7 was detectable by both RNAseq and HCR-ISH and considered a marker of the circadian clock by the authors.

The authors claim that the study reveals insights into the evolutionary origin of circadian timing; Clock is conserved across distant groups of organisms, having a function as a positive regulator of the transcriptional translational feedback loop at the heart of daily timing, but is not a central element of the core feedback loop circadian system in this basal species. Their behavioural and transcriptomic data largely support the claims that Clock is necessary for endogenous daily activity but that the putative molecular circadian system is not self-sustained under constant darkness (this was known already for WT animals)- rather it is responsive to light cycles with altered dynamics in Clock-/- specimens in some core genes under LD. In the main, I think the authors achieved their aims and the manuscript is a solid piece of important work. The Clock-/- animal is a useful resource for examining time-keeping in a basal metazoan.

The work described builds on other transcriptomic-based works on cnidaria, including Nematostellar, and does probe into the molecular underpinnings with a loss-of-function in a gene known to be core in other circadian systems. The field of chronobiology will benefit from the evolutionary aspect of this work and the fact that it highlights the necessity to study a range of non-model species to get a fuller picture of timing systems to better appreciate the development and diversity of clocks.

Strengths:

The generation of a line of Clock mutant Nematostellar is a very useful tool for the chronobiological community and coupled with a growing suite of tools in this species will be an asset. The experiments seem mostly well conceived and executed (NB see 'weaknesses'). The problem tackled is an interesting one and should be an important contribution to the field.

Weaknesses:

I think the claims about shedding light on the evolutionary origin of circadian time maintenance are a little bold. I agree that the data do point to an alternative role for Clock in this animal in light responsiveness, but this doesn't illuminate the evolution of time-keeping more broadly in my view. In addition, these are transcriptomic data and so should be caveated- they only demonstrate the expression of genes and not physiology beyond that. The time-course analysis is weakened by its low resolution, particularly for the RAIN algorithm when 4-hour intervals constrain the analysis. I accept that only 24h rhythms were selected in the analysis from this but, it might be that detail was lost - I think a preferred option would be 2 or 3-hour resolution or 2 full 24h cycles of analysis.

The authors discount the possibility of the observed 12h rhythmicity in Clock-/- animals by exposing them to LD6:6 cycles before free-running them in DD. I suggest that LD cycles are not a particularly robust way to entrain tidal animals as far as we know. Recent papers show inundation/mechanical agitation are more reliable cues (Kwiatkowski ER, et al. Curr Biol. 2023, 2;33(10):1867-1882.e5. doi: 10.1016/j.cub.2023.03.015; Zhang L., et al Curr Biol. 2013, 23;19, 1863-1873 doi.org/10.1016/j.cub.2013.08.038.) and might be more effective in revealing endogenous 12h rhythms in the absence of 24h cues.

Reviewer #2 (Public Review):

This manuscript addresses an important question: what is the role of the gene Clock in the control of circadian rhythms in a very primitive group of animals: Cnidaria. Clock has been found to be essential for circadian rhythms in several animals, but its function outside of Bilaterian animals is unknown. The authors successfully generated a severe loss-of-function mutant in Nematostella. This is an important achievement that should help in understanding the early evolution of circadian clocks. Unfortunately, this study currently suffers from several important weaknesses. In particular, the authors do not present their work in a clear fashion, neither for a general audience nor for more expert readers, and there is a lack of attention to detail. There are also important methodological issues that weaken the study, and I have questions about the robustness of the data and their analysis. I am hoping that the authors will be able to address my concerns, as this work should prove important for the chronobiology field and beyond. I have highlighted below the most important issues, but the manuscript needs editing throughout to be accessible to a broad audience, and referencing could be improved.

Major issues:

1. Why do the authors make the claim in the abstract that CLOCK function is conserved with other animals when their data suggest that it is not essential for circadian rhythms? dCLK is strictly required in Drosophila for circadian rhythms. In mammals, there are two paralogs, CLOCK and NPAS2, but without them, there are no circadian rhythms either. Note also that the recent claim of BMAL1-independent rhythms in mammals by Ray et al., quoted in the discussion to support the idea that rhythms can be observed in the absence of the positive elements of the circadian core clock, had to be corrected substantially, and its main conclusions have been disputed by both Abruzzi et al. and Ness-Cohn et al. This should be mentioned.

2. The discussion of CIPC on line 222 is hard to follow as well. How does mRNA rhythm inform the function of CIPC, and why would it function as a "dampening factor"? Given that it is "the only core clock member included in the Clock-dependent CCGs," (220) more discussion seems warranted. Discussing work done on this protein in mammals and flies might provide more insight.

3. The behavioral arrhythmicity seen with their Clock mutation is really interesting. However, what is shown is only an averaged behavior trace and a single periodogram for the entire population. This leaves open the possibility that individual animals are poorly synchronized with each other, rather than arrhythmic. I also note that in DD there seem to be some residual rhythms, though they do not reach significance. Thus, it is also possible that at least some individual animals retain weak rhythms. The authors should analyze behavioral rhythms in individual animals to determine whether behavioral rhythmicity is really lost. This is important for the solidity of their main conclusions.

4. There is no mention in the results section of the behavior of heterozygotes. Based on supplement figure 2A, there is a clear reduction in amplitude in the heterozygous animals. Perhaps this might be because there is only half a dose of Clock, but perhaps this could be because of a dominant-negative activity of the truncated protein. There is no direct functional evidence to support the claim that the mutant allele is nonfunctional, so it is important to discuss carefully studies in other species that would support this claim, and the heterozygous behavior since it raises the possibility that the mutant allele acts as a dominant negative.

5. I do not understand what the bar graphs in Figure 2E and 3B represent - what does the y-axis label refer to?

6a. I note that RAIN was used, with a p<0.05 cut-off. I believe RAIN is quite generous in calling genes rhythmic, and the p-value cut-off is also quite high. What happens if the stringency is increased, for example with a p<0.01.

b. It would be worth choosing a few genes called rhythmic in different conditions (mutant or wild-type. LD or DD), and using qPCR to validate the RNAseq results. For example, in Figure 3D, Myh7 RNAseq data are shown, and they do not look convincing. I am surprised this would be called a circadian rhythm. In wild-type, the curve seems arrhythmic to me, with three peaks, and a rather large difference between the first and second ZT0 time point. In the Clock mutants, rhythms seem to have a 12hr period, so they should not be called rhythmic according to the material and methods, which says that only ca 24hr period mRNA rhythms were considered rhythmic. Also, the result section does not say anything about Myh7 rhythms. What do they tell us? Why were they presented at all?

1. The authors should explain better why only the genes that are both rhythmic in LD and DD are considered to be clock-controlled genes (CCGs). In theory, any gene rhythmic in DD could be a CCG. However, Leach and Reitzel actually found that most genes in DD1 do not cycle the next day (DD2)? This suggests that most "rhythmic" genes might show a transient change in expression due to prolonged obscurity and/or the stress induced by the absence of a light-dark cycle, rather than being clock controlled. Is this why the authors saw genes rhythmic under both LD and DD as actual CCGs? I would suggest verifying that in DD the phase of the oscillation for each CCG is similar to that in LD. If a gene is just responding to obscurity, it might show an elevated expression at the end of the dark period of LD, and then a high level in the first hours of DD. Such an expression pattern would be very unlikely to be controlled by the circadian clock.

2. Since there are still rhythms in LD in Clock mutants, I wonder whether there is a paralog that could be taking Clock's place, similar to NPAS2 in mammals.

3. I do not follow the point the authors try to make in lines 268-272. The absence of anticipatory behavior in Drosophila Clk mutants results from disruption of the circadian molecular clock, due to the loss of Clk's circadian function. Which light-dependent function of Clock are the authors referring to, then? Also, following this, it should be kept in mind that clock mutant mice have a weakened oscillator. The effect on entrainment is secondary to the weakening of the oscillator, rather than a direct effect on the light input pathway (weaker oscillators have increased response to environmental inputs). The authors thus need to more clearly explain why they think there is a conservation of circadian and photic clock function.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment

The study provides potentially fundamental insight into the function and evolution of daily rhythms. The authors investigate the function of the putative core circadian clock gene Clock in the cnidarian Nematostella vectensis. While it parts still incomplete, the evidence suggests that, in contrast to mice and fruit flies, Clock in this species is important for daily rhythms under constant conditions, but not under a rhythmic light/dark cycle, suggesting that the major role of the circadian oscillator in this species could be a stabilizing function under non-rhythmic environmental conditions.

Public Reviews:

Reviewer #1 (Public Review):

In this nice study, the authors set out to investigate the role of the canonical circadian gene Clock in the rhythmic biology of the basal metazoan Nematostella vectensis, a sea anemone, which might illuminate the evolution of the Clock gene functionality. To achieve their aims the team generated a Clock knockout mutant line (Clock-/- ) by CRISPR/Cas9 gene deletion and subsequent crossing. They then compared wild-type (WT) with Clock-/- animals for locomotor activity and transcriptomic changes over time in constant darkness (DD) and under light/dark cycles to establish these phenotypes under circadian control and those driven by light cycles. In addition, they used Hybridization Chain Reaction-In situ Hybridization (HCR-ISH) to demonstrate the spatial expression of Clock and a putative circadian clocl-controlled gene Myh7 in whole-mounted juvenile anemones.

The authors demonstrate that under LD both WT and Clock-/- animals were behaviourally rhythmic but under DD the mutants lost this rhythmicity, indicating that Clock is necessary for endogenous rhythms in activity. With altered LD regimes (LD6:6) they show also that Clock is light-dependent. RNAseq comparisons of rhythmic gene expression in WT and Clock-/- animals suggest that clock KO has a profound effect on the rhythmic genome, with very little overlap in rhythmic transcripts between the two phenotypes; of the rhythmic genes in both LD and DD in WT animals (220- termed clock-controlled genes, CCGS) 85% were not rhythmic in Clock-/- animals in either light condition. In silico gene ontology (GO) analysis of CCGS reflected process associated with circadian control. Correspondingly, those genes rhythmic in KO animals under DD (here termed neoCCGs) were not rhythmic in WT, lacked upstream E-box motifs associated with circadian regulation, and did not display any GO enrichment terms. 'Core' circadian genes (as identified in previous literature) in WT and Clock-/- animals were only rhythmic under entrainment (LD) conditions whilst Clock-/- displayed altered expression profiles under LD compared to WT. Comparing CCGs with previous studies of cycling genes in Nematostellar, the authors selected a gene from 16 rhythmic transcripts. One of these, Myh7 was detectable by both RNAseq and HCR-ISH and considered a marker of the circadian clock by the authors.

The authors claim that the study reveals insights into the evolutionary origin of circadian timing; Clock is conserved across distant groups of organisms, having a function as a positive regulator of the transcriptional translational feedback loop at the heart of daily timing, but is not a central element of the core feedback loop circadian system in this basal species. Their behavioural and transcriptomic data largely support the claims that Clock is necessary for endogenous daily activity but that the putative molecular circadian system is not self-sustained under constant darkness (this was known already for WT animals)- rather it is responsive to light cycles with altered dynamics in Clock-/- specimens in some core genes under LD. In the main, I think the authors achieved their aims and the manuscript is a solid piece of important work. The Clock-/- animal is a useful resource for examining time-keeping in a basal metazoan.

The work described builds on other transcriptomic-based works on cnidaria, including Nematostellar, and does probe into the molecular underpinnings with a loss-of-function in a gene known to be core in other circadian systems. The field of chronobiology will benefit from the evolutionary aspect of this work and the fact that it highlights the necessity to study a range of non-model species to get a fuller picture of timing systems to better appreciate the development and diversity of clocks.

Strengths:

The generation of a line of Clock mutant Nematostellar is a very useful tool for the chronobiological community and coupled with a growing suite of tools in this species will be an asset. The experiments seem mostly well conceived and executed (NB see 'weaknesses'). The problem tackled is an interesting one and should be an important contribution to the field.

Weaknesses:

I think the claims about shedding light on the evolutionary origin of circadian time maintenance are a little bold. I agree that the data do point to an alternative role for Clock in this animal in light responsiveness, but this doesn't illuminate the evolution of time-keeping more broadly in my view. In addition, these are transcriptomic data and so should be caveated- they only demonstrate the expression of genes and not physiology beyond that. The time-course analysis is weakened by its low resolution, particularly for the RAIN algorithm when 4-hour intervals constrain the analysis. I accept that only 24h rhythms were selected in the analysis from this but, it might be that detail was lost - I think a preferred option would be 2 or 3-hour resolution or 2 full 24h cycles of analysis.

The authors discount the possibility of the observed 12h rhythmicity in Clock-/- animals by exposing them to LD6:6 cycles before free-running them in DD. I suggest that LD cycles are not a particularly robust way to entrain tidal animals as far as we know. Recent papers show inundation/mechanical agitation are more reliable cues (Kwiatkowski ER, et al. Curr Biol. 2023, 2;33(10):1867-1882.e5. doi: 10.1016/j.cub.2023.03.015; Zhang L., et al Curr Biol. 2013, 23;19, 1863-1873 doi.org/10.1016/j.cub.2013.08.038.) and might be more effective in revealing endogenous 12h rhythms in the absence of 24h cues.

Response: We removed the suggestion that we used 6:6h LD to perform tidal entrainment. We generated this ultradian light condition to address the 24h rhythmicity observed in the NvClk1-/- in 12:12h LD.

Reviewer #2 (Public Review):

This manuscript addresses an important question: what is the role of the gene Clock in the control of circadian rhythms in a very primitive group of animals: Cnidaria. Clock has been found to be essential for circadian rhythms in several animals, but its function outside of Bilaterian animals is unknown. The authors successfully generated a severe loss-of-function mutant in Nematostella. This is an important achievement that should help in understanding the early evolution of circadian clocks. Unfortunately, this study currently suffers from several important weaknesses. In particular, the authors do not present their work in a clear fashion, neither for a general audience nor for more expert readers, and there is a lack of attention to detail. There are also important methodological issues that weaken the study, and I have questions about the robustness of the data and their analysis. I am hoping that the authors will be able to address my concerns, as this work should prove important for the chronobiology field and beyond. I have highlighted below the most important issues, but the manuscript needs editing throughout to be accessible to a broad audience, and referencing could be improved.

Major issues:

(1) Why do the authors make the claim in the abstract that CLOCK function is conserved with other animals when their data suggest that it is not essential for circadian rhythms? dCLK is strictly required in Drosophila for circadian rhythms. In mammals, there are two paralogs, CLOCK and NPAS2, but without them, there are no circadian rhythms either. Note also that the recent claim of BMAL1-independent rhythms in mammals by Ray et al., quoted in the discussion to support the idea that rhythms can be observed in the absence of the positive elements of the circadian core clock, had to be corrected substantially, and its main conclusions have been disputed by both Abruzzi et al. and Ness-Cohn et al. This should be mentioned.

Response: According to our Behavioral and Transcriptomic data, CLOCK function is conserved in constant light condition. In LD context, the rhythmicity is maintained probably by the light-response pathway in Nematostella. We modified our rhythmic transcriptomic analysis and considered the context of the contested results by Ray et al., and discussed it in the revised manuscript.

(2) The discussion of CIPC on line 222 is hard to follow as well. How does mRNA rhythm inform the function of CIPC, and why would it function as a "dampening factor"? Given that it is "the only core clock member included in the Clock-dependent CCGs," (220) more discussion seems warranted. Discussing work done on this protein in mammals and flies might provide more insight.

Response: The initial sentence was unclear. Furthermore, since we restricted our rhythmic analysis to genes only found rhythmic with a p<0.01 with RAIN combined with JTK, NvCipc was no longer defined as rhythmic in free running.

(3) The behavioral arrhythmicity seen with their Clock mutation is really interesting. However, what is shown is only an averaged behavior trace and a single periodogram for the entire population. This leaves open the possibility that individual animals are poorly synchronized with each other, rather than arrhythmic. I also note that in DD there seem to be some residual rhythms, though they do not reach significance. Thus, it is also possible that at least some individual animals retain weak rhythms. The authors should analyze behavioral rhythms in individual animals to determine whether behavioral rhythmicity is really lost. This is important for the solidity of their main conclusions.

Response: Fig. 1 has been modified. We have separated the data for WT and NvClk1-/- animals to provide clarity on the average behavior pattern for each genotype. While the LSP analysis on the population average informs us about the synchronization of the population, it is true that it does not provide insight into individual rhythmicity. To address this, we analyzed individuals in all conditions using the Discorhythm website (Carlucci et al., 2019).

In the revised figure, we have included a comparison plot of the acrophase of 24-hour rhythmic animals between genotypes using Cosinor analysis, which is most suitable for acrophase detection. This plot indicates the number of animals detected as significantly rhythmic, providing direct visual input to the reader regarding individual rhythmicity. Additionally, we have added Table 1, which contains the Cosinor period analysis (24 and 12 hours) of individuals for all genotypes and conditions, further enhancing the clarity of our findings.

(4) There is no mention in the results section of the behavior of heterozygotes. Based on supplement figure 2A, there is a clear reduction in amplitude in the heterozygous animals. Perhaps this might be because there is only half a dose of Clock, but perhaps this could be because of a dominant-negative activity of the truncated protein. There is no direct functional evidence to support the claim that the mutant allele is nonfunctional, so it is important to discuss carefully studies in other species that would support this claim, and the heterozygous behavior since it raises the possibility that the mutant allele acts as a dominant negative.

Response: Extended Data Fig.1 modified. We show NvClk1+/- normalized locomotion over time in DD of the population, comparison of individual normalized behavior amplitude, LSP of the average population and individual acrophase of only rhythmic 24h individuals. Indeed, we cannot discriminate Dominant-negative from non-functional allele.

(5) I do not understand what the bar graphs in Figure 2E and 3B represent - what does the y-axis label refer to?

Response: Not relevant to the revised manuscript.

(6a) I note that RAIN was used, with a p<0.05 cut-off. I believe RAIN is quite generous in calling genes rhythmic, and the p-value cut-off is also quite high. What happens if the stringency is increased, for example with a p<0.01.

Response: We acknowledge your concern regarding the stringency of our statistical analysis. To address this, we opted to combine both RAIN and JTK methods and applied a more stringent p-value cut-off of p<0.01.

(6b) It would be worth choosing a few genes called rhythmic in different conditions (mutant or wild-type. LD or DD), and using qPCR to validate the RNAseq results. For example, in Figure 3D, Myh7 RNAseq data are shown, and they do not look convincing. I am surprised this would be called a circadian rhythm. In wild-type, the curve seems arrhythmic to me, with three peaks, and a rather large difference between the first and second ZT0 time point. In the Clock mutants, rhythms seem to have a 12hr period, so they should not be called rhythmic according to the material and methods, which says that only ca 24hr period mRNA rhythms were considered rhythmic. Also, the result section does not say anything about Myh7 rhythms. What do they tell us? Why were they presented at all?

Response: Regarding the suggestion for independent verification of our RNAseq results, we agree that such validation would enhance the robustness of our findings. To address this, we chose to overlap our identified rhythmic genes under WT LD conditions with those from another transcriptomic study that shared similarities in experimental design. Notably, the majority of overlapping rhythmic genes between the studies are candidate pacemaker genes. We believe that this replication of biologically significant rhythmic genes strengthens the validity and reliability of our results (see Extended Data Fig. 2).

Furthermore, we have decided to remove the NvMhc-st (mistakenly named Myh7, only rhythmic in WT DD in the new analysis) as it does not contribute substantively to the revised version of the manuscript.

(7) The authors should explain better why only the genes that are both rhythmic in LD and DD are considered to be clock-controlled genes (CCGs). In theory, any gene rhythmic in DD could be a CCG. However, Leach and Reitzel actually found that most genes in DD1 do not cycle the next day (DD2)? This suggests that most "rhythmic" genes might show a transient change in expression due to prolonged obscurity and/or the stress induced by the absence of a light-dark cycle, rather than being clock controlled. Is this why the authors saw genes rhythmic under both LD and DD as actual CCGs? I would suggest verifying that in DD the phase of the oscillation for each CCG is similar to that in LD. If a gene is just responding to obscurity, it might show an elevated expression at the end of the dark period of LD, and then a high level in the first hours of DD. Such an expression pattern would be very unlikely to be controlled by the circadian clock.

Response: As we modified our transcriptomic analysis, we do no longer analyze LD+DD rhythmic genes, but any genes rhythmic (RAIN and JTK p<0.01) in each condition. As such we end up with four list of genes corresponding to each experimental conditions.

(8) Since there are still rhythms in LD in Clock mutants, I wonder whether there is a paralog that could be taking Clock's place, similar to NPAS2 in mammals.

Response: see response to (1) > The only NPAS2 orthologous identified in Nematostella NPAS3 showed marginally significance (p=0.013) with RAIN in LD WT suggesting a regulation similar to the candidate pacemaker genes. As such we included within our candidate pacemaker genes list.

(9) I do not follow the point the authors try to make in lines 268-272. The absence of anticipatory behavior in Drosophila Clk mutants results from disruption of the circadian molecular clock, due to the loss of Clk's circadian function. Which light-dependent function of Clock are the authors referring to, then? Also, following this, it should be kept in mind that clock mutant mice have a weakened oscillator. The effect on entrainment is secondary to the weakening of the oscillator, rather than a direct effect on the light input pathway (weaker oscillators have increased response to environmental inputs). The authors thus need to more clearly explain why they think there is a conservation of circadian and photic clock function.

Response: Following the changes in our statistical analysis we reframed the discussion and address directly the circadian and the photic clock function (we call it light-response pathway in the manuscript)

Recommendations for the authors:

We suggest the following improvements:

(1) Please undertake a serious effort to make this work more accessible to non-marine chronobiologists. This includes better explanations, and schemes of the animal when images of staining are shown (e.g. Fig.1b) which include the labeling of relevant morphological structures mentioned in the text (like "tentacle endodermis and mesenteries" (line 132)). Similar issues for mentioned life cycle stages like "late planula stage" (line 133), "bisected physa" (line 149).

Response: Fig. 1b, we outlined the animal shaped and added 2 arrows to locate the tentacle endodermis and mesenteries. We replaced the term late planula stage, by larvae. And we rephrased bisected physa by tissue sampling.

Please attend to details. This includes:

Wrong referrals to figures (currently line 151 refers to EDF2- but should be EDF 1 instead, there is a Fig.3f mentioned in the text, but there is no such Fig.).

Response: Fixed

Mentioning of ZTs when the HCR stainings were performed.

Response: Fixed

Fig.1 a shows a rather incomplete and thus potentially confusing phylogenetic tree. Vertebrates have at least two Clk orthologs (NPAS2 and CLK), please include both, use an outgroup, and rout the tree.

Response: Identifying NPAS2 and CLK orthologous in all species added more confusion into the conclusion. However, we followed the suggestion of adding an outgroup using a CLK orthologous sequence identified in the sponge Amphimedon queenslandica and rout the tree. Thank for the suggestion.

What do the y-axis labels in Figure 2E and 3B refer to exactly? Y-axis label annotations in Fig.3a,d are entirely missing- what do the numbers refer to?

Response: not relevant in the revised manuscript

Fig.2D- is the Go term enrichment referring to LD or DD?

Response: to DD. We made it cleared on the figure 5.

Wording: "Clock regulates genetic pathways." What is meant by "genetic pathways"? There are no "non-genetic pathways". Could one simply say: "Clock regulates a variety of transcripts".

Response: We modified our threshold to use only p.adj<0.01, which reduced the GO term numbers. We removed “genetic pathways” and now address the specific pathways: cell-cycle and neuronal.

The use of the term "epistatic" is confusing (line 219), i.e. that light is epistatic to Clock. In genetics, epistasis is defined as the effect of gene interactions on phenotypes. To a geneticist, this implies that there is a second gene impacting on the phenotype of the Clock mutants. Please re-word.

Response: “light is epistatic on Clock” has been re-phrased.

The provided Supplementary tables are not well annotated. Several of them need guess-work about what is shown. For instance, for Supplementary Table 1, the Ns are unclear, which in total can go up to almost 200 per condition-genotype, but only about 30 animals for each were tested. Thus, where do the high totals in the LSP table come from? What do the numbers of each periodicity mean? Initially one might assume it was the number of animals that showed a periodogram peak at a given periodicity, but it seems that cannot be. Maybe it counted any period bin over statistical significance? Please clarify with better descriptions and labels.

Response: Supplementary tables are now clearly annotated on their first Tabs. About Fig.1, we already addressed this point in the public review.

Albeit not essential, it would be more reader-friendly to also add a summary table with average period and SD, power and SD, and percentage rhythmicity to the main figure.

Response: Table 1 is added: it contains individual count of rhythmic animals (24h and 12h) with Cosinor. However, using Discorhythm we had to ask for a specific Period. Thus, we can only provide animal count significant for a given period value. And not an estimation of their own period.

(2) Some of the terminology is quite confusing, in particular the double meaning of the word "clock" (i.e the pacemaker and the transcription factor). This is not a specific problem to this manuscript, but it would be helpful for the readability to try to improve this.

Could the gene/transcript/protein be spelled: clk and Clk?

Alternatively, for clarity- how about talking about "core pacemaker genes," "CLOCK-dependent rhythmic genes" and "CLOCK-independent rhythmic genes"?

Response:

Clock/CLOCK > NvClk / NvCLK and the mutant is NvClk1-/-

Core clock genes > candidate pacemaker genes.

CLOCK-dependent CCG > this notion no longer exists in the revised manuscript.

CLOCK-independent CCG > this notion no longer exists in the revised manuscript.

(3) The dismissal of the 12h rhythmicity in Clock-/- animals is not really convincing and should be reconsidered. LD6:6 cycles (before free-running animals in DD) is likely a not particularly robust way to entrain tidal animals. Recent papers show inundation/mechanical agitation are more reliable cues (Kwiatkowski ER, et al. Curr Biol. 2023, 2;33(10):1867-1882.e5. doi: 10.1016/j.cub.2023.03.015; Zhang L., et al Curr Biol. 2013, 23;19, 1863-1873 doi.org/10.1016/j.cub.2013.08.038.) and might be more effective in revealing endogenous 12h rhythms in the absence of 24h cues.

Response: We removed the proposition of using 6:6hLD as Tidal entrainment. Instead, the LD 6:6 experiment reveals the direct light-dependency of the NvClk1-/- mutant.

(4) There are significant questions raised on the validity of BMAL1-independent rhythms in mammals as suggested by the Ray et al study. See DOI: 10.1126/science.abe9230 and DOI: 10.1126/science.abf0922

These technical comments should also be taken into account and the discussion adjusted accordingly to better reflect the ongoing discussions in the chronobiology field.

Response: We modified our rhythmic analysis. As we cannot use BHQ or adjusted p-value which resulted in very genes, we defined 24h-rhythmic genes if p<0.01 with two different algorithms (RAIN and JTK). We propose this compromise to reduce the risk of false-positive. Furthermore, we discussed our methodology in the light of the significant questions raised by these papers you cited. We thank the reviewer for this important point.

(5) The HCR stainings for clk are not very convincing. Normally, HCR should have more dots. In principle, the logic of HCR is such that it detects individual mRNA molecules in the cell. Thus, having only one strong dot/cell like in Fig.1b doesn't make much sense.

Response: We were the first surprised by this single dot signal. We are experienced users of HCRv.3 across different species. We decided to remove the close-up (for further investigations) but to keep the full animal signal. According to our approach it is a convincing signal. However, the doty nature of the signal itself it is not easy to make it highly visible at full scale animal on the picture. We did our best to show the mRNA signal visible without altering the pattern.

Furthermore, the controls for the HCR in situ hybridization are unclear. In the methods, there are two Clock probes described (B3 & B5) and two control probes (B1 & B3), however, in the negative control image, a combination of one Clock (B1) and one control (B3) probes is used and is unclear what "redundant detection" means in the legend of figure S2.

Response: Considering the nature of the signal (single of few dots), we decided to use two probes with 2 different fluorophores. A noise is by nature random. Our hypothesis was: only overlapping fluorescent dots are true signal of NvClk mRNA.

For Control probes we used two zebrafish probes labelling hypothalamic peptides.

Based on the experience with non-Drosophila, non-mouse animal model systems the reviewers assume that non-sense mediated mRNA decay (NMD) is not strongly initiated upon Crispr-induced premature STOP-codons. If this assumption is correct it would be worth to mention it. Alternatively, it would be worth testing if Nematostella induces NMD, as this would be a great control for the HCR and the mutation itself. At which ZT was the HCR done?

Response: We performed the HCR at ZT10 when NvClk is described to be at peak. It is now indicated in the Fig. 1b. The RNAseq detected a higher quantity of NvClk1 mRNA in the NvClk1-/- (see Fig. 4a). mRNA quantity regulation involves transcription, stabilization, and degradation. At this stage, we cannot identify which specific step is affected.

For Fig.1c- please provide the binding site and sequence in the figure, simply include EDF 1 in the main figure.

Response: We generated a clear indication in the new Fig.1c and EDF. 1b about the protein domains, the CRISPR binding site and the consequences on the DNA and AA sequences.

(6) Please provide the individual trace data for the behavioral analyses either as supplementary files or as a link to an openly accessible database like DRYAD (see also comment 7 in the public review of reviewer 2). Maybe this is what is shown in Supplementary Table 1, but it is really not clear what is actually shown.

Response: Fig.1 is updated. Table 1 is added. Supplementary Table 1 contains individual normalized locomotor data of each polyps for each genotypes and light conditions. Supplementary Table 2 contains the cosinor individual rhythmic behavior analysis based on the Supplementary Table 1.

(7) It is not really clear if the mutation is a true loss-of-function or could also be dominant negative. While this is raised in the discussion, it should be more carefully considered. The reason why a dominant negative would be unlikely is unclear. More specifically also see comment 8) in the public review of reviewer 2.

Response: Indeed, the results cannot tell us if it is a true loss of function, a dominant negative or non-functional allele. We addressed it in the first part of the discussion.

(8) The pretty small overlap of rhythmic transcripts in LD and DD could reflect the true biology of a more core clock driven-process under constant conditions and a more light-driven process under LD. But still- wouldn't one expect that similar processes should be rhythmic? If not, why not?

It would certainly add strength to the data if for one or two transcripts these results were independently verified by qPCR from an independent sampling. This could even be done for just two time points with the most extreme differences.

Response: We appreciate the reviewer's comments and concerns regarding the overlap of rhythmic transcripts in different conditions. In response to the reviewer's query, we revised our interpretation of the transcriptomic data, acknowledging the limited overlap between light and genotype conditions in our study. This prompted us to reconsider the underlying biological processes driving rhythmic gene expression under constant conditions versus light-dark cycles.

Regarding the suggestion for independent verification of our RNAseq results, we agree that such validation would enhance the robustness of our findings. To address this, we chose to overlap our identified rhythmic genes under WT LD conditions with those from another transcriptomic study that shared similarities in experimental design. Notably, the majority of overlapping rhythmic genes between the studies are candidate pacemaker genes. We believe that this replication of biologically significant rhythmic genes strengthens the validity and reliability of our results (see Extended Data Fig. 2).

(9) Expression of myh7 : Checking for co-expression should be pretty straightforward by HCR. This is what this type of staining technique is really good for. Please do clk and myh7 co-staining if you want to claim co-expression. Otherwise don't make such a claim.

Response: We agree that checking for co-expression should be straightforward by HCR. However, due to time constraints during the revision period, we are unable to conduct the double in-situ experiment. Additionally, upon careful consideration, we recognize that including myhc-st (mistakenly named myh7) staining and co-expression analysis would not significantly contribute to the main conclusions of our study. Therefore, we have decided to remove this analysis from the revised manuscript.

(10) Missing methodological details:

The false discovery rate for each analysis should be included (see Hughes et al.,: "Guidelines for Genome-Scale Analysis of Biological Rhythms," 2017).

Response: THE FDR is indicated for each gene in supplementary table 3

Fig.1f- continuous light- please provide a spectrum (If there is no good spectrophotometer available, please provide at least manufacturer information.

Response: Unfortunately, we don’t have a good spectrophotometer available during the time of the revision. We added to the method the reference of the lamp. We found the light spectrum provided by the supplier. However, we did not add it to the revised manuscript.

Spectrum of the Aquastar t8

Also, it would be easier for the reader, if the measurements of light intensity are provided in photons, because this is what the light receptors ultimately measure.

Response: Modified.

Fig.2E- please add the consensus sequence used for circadian E-box vs. E-box to the figure.

Response: In the revised manuscript Fig.4c, we show which E-box motifs we extracted for our promoter analysis. We as well changed our analysis and did no longer use HOMER, but we directly extracted promoter sequences and looked for canonical Ebox CANNTG and Circadian Ebox CACGTG and generate a Circadian Ebox enrichment output per gene promoter.

(11) There has been some discussion about the evolutionary statement as stated by the authors. It appears that depending on the background of the reader, this can be misunderstood. We thus suggest to more clearly point out where the author thinks there is evolutionary conservation (a function for clk in the circadian oscillator under constant light or dark conditions) versus where there is no apparent evolutionary conservation (the situation under light-dark conditions).

Response: In the revised manuscript we proposed a conserved function of NvCLK in constant darkness, and a light-response pathway compensating in LD conditions in the mutant.

Please also consider the major comments 8 and 9 of the common review from reviewer 2.

Reviewer #1 (Recommendations For The Authors):

The hybridization chain-reaction ISH is OK but, I'm not sure I understand the control condition-this should be clarified. I would also welcome the use of Clock-/- animals in HCR as another, more direct level of control. In addition, the authors state that the Myh7 probes hybridise in anatomical regions resembling those for Clock (Fig 3e). It would be better to duplex these two probe sets with different fluors for a better representation of the relative spatial distributions of each transcript.

Response: We agree that checking for co-expression should be straightforward by HCR. However, due to time constraints during the revision period, we are unable to conduct the double in-situ experiment. Additionally, upon careful consideration, we recognize that including myhc-st (mistakenly named myh7) staining and co-expression analysis would not significantly contribute to the main conclusions of our study. Therefore, we have decided to remove this analysis from the revised manuscript.

We clarified in the methods the control probes design.

Minor points:

Figure legends do not all convey sufficient detail. For instance, Figure 1c needs a better explanation. Figure 3e- are these images both WT? Fig 3f doesn't exist and other figure text references do not align with figures and need an overhaul.

Response: All errors have been fixed.

Reviewer #2 (Recommendations For The Authors):

Major issues:

(1) The authors need to introduce their model system better for a broad audience. What are the tissues/cells that express Clock at a higher level? What is their function, does this provide a potential explanation for their specific Clock expression, and how CLOCK might regulate behavior? Terms such as "tentacle endodermis and mesenteries" (line 132), "late planula stage" (line 133), "bisected physa" (line 149) would need some explanation.

Response: We modified term such as planula to larvae, and bisected physa to tissue samples.

1. Some of the terminology used is quite confusing, because of the double-meaning of the word "clock" (i.e the pacemaker and the transcription factor). The authors use terms such as "clock-controlled genes", "core clock genes", "CLOCK-dependent clock-controlled genes", "neo-clock-controlled genes". Is there any way to help the reader? Here are several suggestions: "core pacemaker genes," "CLOCK-dependent rhythmic genes" and "CLOCK-independent rhythmic genes".

Response: all the terminology has been clarified, see previous comments

1. Also in the abstract, there is mention of "hierarchal light- and Clock-signaling" (52-3) - is this related to the statement on line 219 that light is epistatic to Clock? I do not quite understand what epistatic would mean here. Who is upstream of whom? LD modifies rhythmicity in Clock mutant animals, but Clock mutations also impact rhythmicity in LD. Also, as epistasis is defined as the effect of gene interactions on phenotypes - what is the secondary gene impacting the phenotype of the Clock mutants? I am not sure the term epistatic is appropriate in the present context.

Response: Indeed, Epistatic is a genetic term which might be unclear in this context. We removed it.

1. The control for the in situ hybridization is unclear. In the methods, there are two Clock probes described (B3 & B5) and two control probes (B1 & B3), however, in the negative control image, a combination of one Clock (B1) and one control (B3) probe is used, I am not sure what "redundant detection" means in the legend of figure S2. Also, the sequences of each Clock probe should be provided. It might be worth testing the Clock mutant the authors generated. Clock mRNA could be reduced due to non-sense, mediated RNA decay, since the mutation causes a premature stop codon. This would be a great additional control for the in situ hybridization. Even better would be if, by chance, the probes target the mutated sequence. The signal should then be completely lost.

Response: HCR is a tilling probe. Which means the target transcript is covered by dozens of successive DNA sequence “primer-like” which allow the HCRv.3 technology. We cannot design a mutant probe specific with this technology.

(5) I have concerns with rhythmic-expression calls, particularly as there is so little overlap between LD and DD, and that a completely different set of rhythmic genes is observed in Clock mutant and wild-type animals. I am not an expert in whole-genome expression studies, so I hope one of my colleague reviewers can weigh in.

When describing rhythmicity analysis in the Methods, it states that Benjamini-Hochberg corrections were applied to account for multiple comparisons. However, the false discovery rate for each analysis should be included (see Hughes et al.,: "Guidelines for Genome-Scale Analysis of Biological Rhythms," 2017).

Response: As explained before we cannot used Benjamini-Hochberg corrections as only few genes (mostly oscillator gene pass the threshold). As such we combined two different algorithms (RAIN and JTK) with a p<0.01 to detect confidently rhythmic genes while reducing the risk of false-positives.

Minor issues:

(1) Environmental inputs are not "circadian", as written in the title.

Response: Title modified

(2) In the abstract, the description of the Clock mutant behavioral phenotypes is hard to follow, with no mention of whether or not Clock mutant animals are behaviorally rhythmic or arrhythmic in constant conditions.

Response: corrected

(3) Abstract: A 6/6 h LD cycle is not a compressed tidal cycle as written in the abstract. Light is not an input to tidal rhythms.

Response: corrected

(4) Line 101: timeout is not a core clock gene in animals.

Response: we removed it from the candidate pacemaker genes.

(5) What is the evidence for the role of PAR-Zip proteins in the Nematostella clock? The reference provided does not mention those.

Response: There is no functional data in Nematostella yet to support their role within the pacemaker. However based on their rhythmicity in LD and protein conservation, we included them within the candidate pacemaker genes list. The refences have been corrected.

(6) Line 125. should refer to Fig 1C when describing the Clock protein.

Response: corrected

(7) Line 143-4. based on the figure, the region targeted by gRNA was not "close to the 5' end" as stated, it is closer to the middle of the gene sequence as shown in Figure 1C. A more accurate description would be a region in between the PAS domains.

Response: Indeed we modified the figure and the text.

(8) Line 150. The mutant allele is described as Clock1 initially, then for the rest of the paper as Clock-. SInce it is not clear that the allele is a null (see major comment #8), Clock1 should be used throughout the manuscript.

Response: the allele is named NvClk1 in the revised manuscript

(9) Figure 2A, the second CT/ZT0 is misplaced.

Response: Fig. 2 modified in the revised manuscript

(10) Figure legend for 2E and 3B. "The 1000bp upstream ATG" is unclear. I guess it means that 1000bp upstream of the putative initiation codon was used.

Response: Right, and in the revised version we analyzed 5kb upstream the putative ATG.

(11) Line 164. The authors write "We discovered..." , but wasn't it already known that these animals are behaviorally rhythmic?

Response: Fixed

(12) It would be worth mentioning in the results section the reduced amplitude of rhythms in LL compared to DD (in WT and seemingly also in Clock mutants).

Response: Indeed, we observed a significant reduction in the mean amplitude in the NvClk1-/- in DD and LL compared WT and NvClk1-/- in LD, DD and LL. However, as rhythmicity is lost by virtually all mutants in LL and DD we do not think these results add to the current interpretation of the gene function.

(13) Please correct the figure numbers in the main text, there are several mistakes.

Response: Done

(14) Line 196, most genes in the quoted study did not cycle on day 2, so whether they are truly clock controlled is questionable.

Response: We agree, identifying free-running cycling genes in cnidarian remains a challenge to overcome. One of the limitations of this study was to detect rhythmic genes in LD which conserved rhythmicity in DD. However, considering different transcriptomic studies (cited in the discussion) it seems that in the cnidaria phyla rhythmic genes in LD are not necessarily the one we identified rhythmic in DD.

(15) Line 204-206 needs to be rephrased. It is confusing.

Response: rephrased

(16) Line 216. Rephrase to something like: "A similar finding was made for."

Response: rephrased

(17) "Clock regulates genetic pathways" sounds quite odd. Do you mean it regulates preferentially specific genetic (or maybe better, molecular) pathways?

Response: rephrased

(18) Figure 4 and legend: Dashed lines indicating threshold are missing. Do the black and red dots represent WT and Clock-/-, as indicated in the legend, or up/down, as indicated in the figures?

Response: Fig.5 modified accordingly. Colors in the Volcano plot indicate Up- (black) versus Down- (red) regulated. It is now coherent within the figure.

(19) Legend for Extended figure 1. "Immature peptide sequence" is incorrect.

Response: rephrased

(20) Extended data Figure 4. What the asterisks labels is unclear.

Response: EDF4 was modified and become EDF2 with different content. The * indicates NvClk mRNA

(21) Line 228. Gene "isoforms". I guess the authors mean "paralogs".

Response: corrected.

(22) Line 232-3/Figure 3e. Please include a comparable image of the Clk ISH to facilitate the comparison of the spatial expression pattern. In addition, where and what is the "analysis" referred to - "the spatial expression pattern of Myh7 closely resembled that of Clock, as evidenced by our analysis"?

Response: the analysis has been removed from the revised manuscript because we currently cannot perform the double ish.

(23) Line 282-3. As mentioned above, it is difficult to be sure that circadian behavior is lost, if only looking at a population of animals.

Response: Fig.1 corrected

(24) Line 301-5. Rephrase.

Response: Rephrased

(25) Line 325. I am not convinced that the author can say that their mutant is amorphic. See Major comment 8.

Response: corrected.

(26) Line 351 "simplifying interactions with the environment". Please explain what is meant here.

Response: this confusing sentence has been removed from the revised manuscript

Version 2

Abstract

Abstract

The circadian clock enables anticipation of the day/night cycle in animals ranging from cnidarians to mammals. Circadian rhythms are generated through a transcription-translation feedback loop (TTFL or pacemaker) with CLOCK as a conserved positive factor in animals. However, the functional evolutionary origin and mechanism of action of CLOCK in basal animals are unknown. In the cnidarian Nematostella vectensis, pacemaker genes transcription including NvClk (the Clock ortholog) appears arrhythmic under constant light conditions, questioning the role of NvCLK. Utilizing CRISPR/Cas9, we generated a NvClk allele mutant (NvClk1), revealing circadian behavior loss in constant light conditions (LL and DD) while a 24-hour rhythm was maintained under light-dark condition (LD). Transcriptomics showed distinct rhythmic genes in wild-type (WT) genes in LD compared to DD. The LD NvClk1-/- showed comparable numbers of rhythmic genes, whereas they were greatly reduced in DD. Furthermore, the LD NvClk1-/- showed alterations of temporal pacemaker genes expression, affecting their potential interactions. Additionally, differential expression of non-rhythmic genes associated with cell division and neuronal differentiation was observed. These findings suggest that while the light-responsive pathway can partially compensate for circadian clock disruption, the Clock gene has evolved in cnidarians to maintain 24-hour rhythmic physiology and behavior in constant conditions.

Introduction

Throughout the history of life on Earth, organisms have had to adapt to a constantly changing environment, including the ∼24-hour daily rhythm of light/dark, driving the development of endogenous biological clocks. The circadian clock, which is entrained by external stimuli such as light, enables the organism to anticipate the onset of the light and dark phases and synchronize its physiology and behavior in harmony with the environment. This, in turn, enhances the organism’s fitness and survival1–3. From single-celled organisms to metazoans, circadian clocks have evolved multiple times, highlighting their importance to living organisms1,2. Despite the fundamental role circadian clocks play in regulating the rhythmicity of living organisms, their evolutionary origin and intricate molecular mechanisms remain ambiguous in early diverging animal lineages, such as cnidaria.

Rhythmic phenomena, including calcification, reproduction, and diel behavior patterns, have been examined in cnidarian species3–8. While some of these patterns were found to be directly triggered by environmental stimuli, such as light, others persist in the absence of external cues, suggesting the presence of an internally generated and self-sustaining circadian clock6,9,10. At the molecular level, cnidarians possess homologs of putative core pacemaker genes found in bilaterians11–13. Several studies have shown that most of these genes display diel expression patterns under light/dark cycles. However, in contrast to most animals, their oscillation generally ceases without the presence of light cues14–16. Thus, how the core pacemaker genes orchestrate rhythmic gene expression and circadian behaviors in cnidarians remains unclear.

One of the most studied cnidarian species in the field of chronobiology is the estuarine sea anemone, Nematostella vectensis. Few studies have shown that in diel lighting, the locomotor behavior of Nematostella has a ∼24-h rhythm that is maintained under constant conditions suggesting it is regulated by an endogenous circadian clock6,17–19. In accordance, the Nematostella genome codes for conserved core pacemaker genes such as NvClk, NvCycle, and the cryptochromes NvCry1a and NvCry1b12,13. The proposed circadian clock model in Nematostella is composed of the positive transcription factors (bHLH-PAS family), NvCLK and NvCYCLE, that heterodimerize and upregulate light-dependent cryptochrome genes in the feedback loop, and NvPAR-bZIPs in the feed-forward loop, which repress the transcription of the positive elements11,12,19. More recently, the NvCLK-interacting pacemaker, NvCIPC was predicted to act as an additional repressor of the NvCLK:NvCYCLE dimer19. However, in contrast to the free-running oscillation demonstrated for Nematostella behavior6,17,19, transcriptional expression profiles of most candidate genes implicated in the pacemaker do not retain their oscillation period in the absence of light14,15,19.

Here, we employed the CRISPR/Cas9-mediated genome editing system to establish a NvClk mutant (NvClk1-/-) Nematostella. By combining behavioral monitoring and transcriptomic analysis, we aimed to elucidate the role of NvClk in regulating rhythmic locomotor activity and gene expression under varying light conditions. Our study revealed a robust light response pathway capable of compensation and a conserved function of CLOCK as a timekeeper in the absence of a light cue.

Results

Phylogenetic analysis and spatial expression pattern of NvClk

Phylogenetic analysis of NvCLK protein sequences revealed that NvClk is positioned within the cnidarian branch (Fig. 1a). It contains a basic helix–loop–helix (bHLH) DNA binding domain and two Per-Arnt-Sim (PAS) domains, similar to the protein structure found in other animals. PAS domains are crucial structural motifs involved in protein-protein interactions that drive the self-sustaining molecular mechanism underlying the circadian clock20,21.

NvClk1-/- cannot maintain circadian behavior in non-diel light conditions.

(a) Phylogenetic tree showing the evolutionary relationship of CLK orthologs across different animal species. (b) In situ hybridization of NvClk in the WT juvenile with scale bars representing 0.1mm. (c) Schematic representation of the NvClk gene in grey, with the open reading frame (ORF) in dark grey and the conserved protein domains bHLH (yellow) and PAS1 and PAS2 (dark red). The CRISPR generated NvClk1 allele has a +20nt insertion after the PAS1 domain, represented by a black arrowhead. NvCLK dimerizes via its three functional domain with NvCYCLE binding the CACGTG ebox to drive rhythmic transcription. (d-g-j-m). Normalized Movement (a.u), hourly binned over 72h, under different light conditions: 12h light:12h dark, continuous dark (Dark - Dark), continuous light (Light - Light), and 6h light:6h dark. The black line represents the WT and the red line represents the NvClk1-/- mutant. (e-h-k-n) Lomb-Scargle Periodograms for each corresponding light condition. The significant period value (p<0.01) is indicated for each genotype in the top left corner of each graph. (f-i-l-o) Phase detection (Cosinor) and genotype comparison of 24h rhythmic individuals. See number nrhythmic/ntotal on the x-axis indicating for each genotype the number of 24h-rhythmic animals over the total population.

To localize NvClk expression at the polyp stage, in situ hybridization chain reaction (HCRv.3) was performed. Polyps were sampled at ZT10, i.e., peak expression of NvClk6,12,22. NvClk specific expression was observed throughout the animal tissue, and enriched expression was visible in the tentacle endodermis and mesenteries, while no signal was observed in the negative control (Fig.1b, Extended Data Fig.1a). This expression pattern resembled the expression observed at the larvae stage22. To date, functional manipulation of the NvClk gene has not been performed in basal animal lineages including cnidarians, and its function is unknown in cnidarians20,21 (Fig.1a).

Generation of NvClk1-/- Nematostella

To investigate the function of NvClk in Nematostella, we employed the CRISPR-Cas9 system to generate mutants. Based on existing knowledge from mouse and Drosophila models, we hypothesized that NvCLK:NvCYCLE dimer binds to the DNA motif CACGTG within the promoter of rhythmic target genes (Fig. 1c). Guide RNA (gRNA) was synthesized to target a region between the two PAS domains of the NvClk coding sequence (CDS). This gRNA, along with the Cas9 endonuclease, was microinjected into zygotes (Methods). Subsequently, F0 animals were outcrossed with wild type (WT), and the F1 progeny were raised to adulthood. Genotyping of F1 polyps identified 10 different mutated alleles, with six displaying a frame-shift mutation, including one with a 20 bp insertion (NvClk1), resulting in a premature stop codon (Extended Data Fig. 1b). The predicted 203 amino acid truncated protein lacked 459 amino acids, including one co-factor dimerization PAS domain (Fig. 1c, Extended Data Fig. 1b). To obtain homozygous NvClk1-/- polyps, we crossed heterozygous NvClk1 F1 animals. Genotyping of F2 polyps confirmed the expected 25% frequency of NvClk1-/- mutants. Subsequently, we intercrossed NvClk1-/- animals to obtain F3 NvClk1-/- polyps for use in subsequent experiments aimed at assessing the role of NvClk in regulating behavioral and genetic rhythms.

NvClk is necessary to maintain circadian behavior in constant conditions

To assess the impact of the NvClk1 mutation on circadian rhythm, we monitored locomotor behavior of WT and NvClk1-/- polyps under different light conditions (Supp. Table 1). Both the WT and NvClk1-/- populations exhibited a 24-hour periodicity in 12:12 hour Light-Dark (LD) cycles (Fig. 1d,e), with 15 out of 21 WT animals displaying 24-hour rhythmicity compared to only 9 out of 20 NvClk1-/- animals (Fig. 1f, Table 1, Supp. Table 2). The average acrophase for WT polyps (13.3 hours) was significantly lower than for NvClk1-/- polyps (17.3 hours) (Fig. 1f, Table 1).

We then investigated locomotor behavior under continuous conditions, namely continuous dark (DD) or continuous light (LL). WT polyps exhibited a 22-hour rhythmic behavior under both constant light conditions, with 17 out of 25 WT polyps displaying a 24-hour rhythm under DD and 7 out of 25 under LL (Fig. 1g-l). In contrast, few NvClk1-/- polyps displayed rhythmic behavior under constant conditions (1 out of 24 in DD and 1 out of 26 in LL) (Table 1). Additionally, we observed an intermediate phenotype in the locomotor behavior of heterozygous polyps for the NvClk1 allele in DD (Extended Data Fig. 1c-f).

To further explore the 24-hour rhythm of NvClk1-/- polyps under LD, we tracked locomotor activity under a 6-hour light: 6-hour dark (LD 6:6) cycle after a regular diel 72-hour entrainment under 12:12 LD. While WT polyps maintained a marginally significant periodicity of 22 hours, NvClk1-/- polyps displayed a 12-hour rhythm at the population level (Fig. 1m-o). Specifically, we identified an important difference of 12- hour rhythmic individual polyps between WT and NvClk1-/- groups (1 out of 25 WT polyps vs. 13 out of 26 NvClk1-/- polyps) (Table 1). Notably, entrainment with LD 6:6 did not lead to a 12-hour rhythm in DD for both WT and NvClk1-/- polyps (Extended Data Fig. 1g-i).

NvClk regulates rhythmic gene expression differentially in response to light conditions

To investigate the underlying molecular correlates of the behavioral phenotype found in NvClk1-/- polyps, we conducted transcriptional profiling. WT and NvClk1-/- polyps were sampled seven times every 4 hours over 24 hours under LD and DD conditions (Fig. 2a). We defined true rhythmic genes using a combination of both RAIN (rhythmicity analysis incorporating non-parametric methods) and JTK_CYCLE algorithms (Fig. 2b, Supp. Table 3). Within our dataset, we identified 119 genes rhythmic under LD and 107 rhythmic genes under DD in WT polyps (Fig. 2b). In NvClk1-/- polyps, we detected 147 rhythmic genes under LD and only 37 under DD (Fig. 2b).

NvClk1-/- shows rhythmic genes reduction in constant darkness with altered rhythmic features.

(a) Overview of the experimental design used to generate RNA-seq data. Polyps were entrained for 72 hours before sampling at 4-hour intervals over a 24-hour period (dark arrows) in both LD and DD cycles. (b) Venn diagram comparing the total number of 24h rhythmic genes identified in WT and NvClk1-/- in LD and DD cycles with a p <0.01 with RAIN and JTK. (c) Average acrophase comparison between rhythmic genes in LD and DD in WT polyps. Mann-Whitney test, p<0.001. (d) Average acrophase comparison between rhythmic genes in LD and DD in NvClk1- polyps. Mann-Whitney test, p>0.05. (3) Average relative amplitude comparison between rhythmic genes in LD and DD in WT polyps. Mann-Whitney test, p<0.05. (d) Average relative amplitude comparison between rhythmic genes in LD and DD in NvClk1-/- polyps. Mann-Whitney test, p>0.05. (c-f) sample size (n) indicated below each boxplots.

The rhythmic genes in WT polyps displayed a higher acrophase under DD compared to LD (17.20h vs 12.93h, Fig. 2c). However, no differences were detected between LD and DD rhythmic genes in NvClk1-/- polyps (Fig. 2d). Similarly, the relative amplitude (rAMP) of DD rhythmic genes was higher in WT polyps compared to LD (0.61 vs 0.43, Fig. 2e), but no rAMP difference was observed between LD and DD rhythmic genes in NvClk1-/- polyps (Fig. 2f).

We performed a clustering analysis on the rhythmic genes using the DPGP model (Dirichlet process Gaussian process mixture model). The average number of genes per cluster between LD and DD conditions in WT polyps did not differ significantly (7.3 vs 7.6, Fig. 3a, Supp. Table 4). However, in NvClk1-/- polyps, the average number of genes per cluster was significantly lower in DD compared to LD (4.1 vs 8.6, Fig. 3b). Specifically, in WT DD polyps, we observed clusters with high numbers of genes peaking at subjective night (Fig. 3c, Supp. Table 4). We did not identify any GO term enrichment to any of the clusters.

NvClk1-/- shows temporal gene clusters recruitment alteration in constant darkness.

(a) Gene per cluster count comparison between LD and DD in WT polyps. Number of cluster indicated below each boxplot. Mann-Whitney, p > 0.05. (b) Gene per cluster comparison between LD and DD in NvClk1-/-. Sample size (n) indicated below each boxplot. Mann-Whitney, p < 0.05. (c) Representative temporal gene cluster for each condition. For both light conditions, from top to bottom cluster peaking at day, then twilight, then at night. Number of gene for each specific cluster indicated on the x-axis.

NvClk regulates temporal expression pattern of pacemaker genes

In line with previous findings in Nematostella12,14, candidate pacemaker genes showed arrhythmic expression under DD conditions (Fig. 4a, Supp. Table 3). However, the altered expression patterns observed in LD NvClk1-/- polyps, compared to LD WT polyps (Fig. 4a, Supp. Table 3), suggest a complex regulatory effect.

NvClk1-/- alters temporal pacemaker genes expression.

(a) Four pacemaker gene are individually plotted showing the read count over 24h in LD and DD in WT (black) and NvClk1-/- (red). The continuous line represents significant rhythmicity (RAIN&JTK p<0.01) while the dashed line indicates no rhythmicity. (b) correlation matrix of candidate pacemaker genes expression in LD for WT on the left and NvClk1-/- on the right. (c) schematic representation of the promoter sequences analyses 5kb upstream the putative ATG. Black boxes represent canonical E-boxes while circadian E-boxes are green. Below the logo motif we used to identify canonical and circadian Ebox. (d) Circadian / Canonical ratio (in %) per condition. Kruskal-Wallis, multiple comparison, a vs b : p<0.05.

To systematically analyze the mutation’s impact on these genes, we constructed a correlation matrix based on their expression levels. In LD WT polyps, genes clustered into two groups: one with NvClk peaking during the day, anticorrelated to the second cluster (NvPar-bzipc and NvCipc) peaking at night. In LD NvClk1-/- polyps, the second cluster contained two additional genes and showed weaker anticorrelation with the NvClk cluster (Fig. 4b).

To go further into the regulatory mechanisms downstream of the pacemaker, we examined the presence of circadian E-box motifs (CACGTG) within 5kb upstream of the predicted ATG of rhythmic genes. We calculated circadian/canonical E-box enrichment to account for the total variation in the number of canonical E-boxes (Fig. 4c). Notably, only the candidate pacemaker genes exhibited a significant enrichment in circadian E-boxes in their promoters (15.9%) compared to the WT (5.6%), NvClk1-/- (4.8%) rhythmic genes, and non-rhythmic genes (6.8%) (Fig. 4d).

NvClk coordinates cell division and neuronal pathways in constant darkness

In addition to the transcriptomic rhythmic analysis, our aim was to identify processes regulated by NvClk that may not necessarily exhibit rhythmicity. To achieve this, we conducted a differential gene expression analysis on the total transcriptome between genotypes under each light condition. Under LD conditions, NvClk1-/- polyps exhibited 457 down-regulated genes and 646 up-regulated genes, with no significant enrichment in GO terms observed (Fig. 5a, Supp. Table 4 and 5). However, in DD conditions, NvClk1-/- displayed 2450 down-regulated genes and 1770 up-regulated genes (Fig. 5b, Supp. Table 4). Notably, the down-regulated genes were predominantly enriched in processes related to mitosis, microtubules, and ciliary/flagellar motility. Conversely, the up-regulated genes showed significant enrichment in processes such as the modulation of another organism’s processes, axonal guidance, and sensory perception (Fig. 5b, Supp. Table 5).

NvClk1-/- disrupts cell-cycle and neuronal pathways in constant darkness.

(a) Volcano plot showing the differential expression of genes (DEG) between WT and NvClk1-/- in LD (Left) and DD (right). Dashed line indicate the threshold used to detect DEG (p.adj<0.01). Red dots indicate down regulated genes and black dots up-regulated genes in NvClk1-/- compare to WT polyps (b) Gene Ontology (GO) terms with with significant fold-enrichment (Bonferroni corrected p-value or p.adjusted <0.01) for the DEG analysis in DD. Down regulated genes in Red while Up regulated genes in Black.

Discussion

Conserved behavioral CLOCK function through animal evolution

Our study provides valuable insights into the evolution of circadian clocks by characterizing the effects of the first Clock mutation in a cnidarian, the sea-anemone Nematostella vectensis. Through our behavioral assays, we have shown that NvClk is essential for maintaining rhythmic locomotor activity in the absence of an entraining light cue. Although the rhythmicity of the NvClk1+/- heterozygote polyps was affected in DD, our results could not discriminate a dominant-negative from a full loss of function to identify the nature of this mutation (Extended Data Fig. 1g-i). Studies in various model organisms further support the importance of CLOCK in regulating circadian locomotion. For instance, homozygous DmClk-/- flies exhibit a loss of circadian locomotion in constant darkness23,24, while the DnClk1a zebrafish mutant displayed a shortened period under the same conditions25. Interestingly, MmClock-/- mice maintain robust circadian locomotion in constant darkness, suggesting potential functional redundancy among mammalian clock genes26–29. Overall, these findings support a conserved role of CLOCK in preserving circadian behavioral rhythms in absence of light cues across the distant Nematostella, flies, zebrafish, and mice.

Moreover, the conservation of a 24-hour locomotion rhythm in LD of the NvClk1-/- polyps with a different mean acrophase, revealed a light-response pathway independent of the circadian circuit, consistent with observations in other animal models23,26,30 (Fig. 1f). NvClk1-/- polyps exposed to a 12:12h LD cycle exhibited a 24-hour period, whereas those exposed to a 6:6h LD cycle displayed a 12-hour period. Notably, nearly no WT polyps exhibited a 12-hour rhythm under this condition, suggesting that the circadian clock overrides the light-response pathway (Fig. 6a). While some of the circadian factors can directly sense the light such as CRY proteins31, 29 putative NvOpsin have been identified in the genome which could be involved in the light-response pathway32. Behavioral tracking of NvClk1-/- polyps exposed to different wavelength could help to identify candidates for further functional studies of the light-response pathway.

Summary of NvClk function in the regulation of Nematostella circadian rhythmicity.

Transcriptional rhythmicity plasticity downstream NvClk

At the transcriptomic level, previous studies in Nematostella have shown large changes in the transcriptional profile of many genes after a single day of constant darkness, including the candidate pacemaker genes that were found arrhythmic, despite sustain circadian locomotion12,14,22. Consistent with previous transcriptomic analysis in cnidarian7,14,16,33, most of rhythmic genes identified in LD were different from those identified in DD in the WT polyps. Notably, they displayed higher mean acrophase and larger mean amplitude in DD suggesting a differential regulation in response to light conditions, which has not been investigated in previous cnidarian studies. Additionally, an overlap between our LD rhythmic genes and Leach et al. dataset14 suggests the robustness of pacemaker rhythmic transcription in LD conditions compared to downstream rhythmic genes (Extended data Fig. 2). Overall, these observations suggest a potential plasticity of the pacemaker in selecting specific genes depending on environmental conditions.

Similar to the rhythmic behavior of NvClk1-/- polyps under LD conditions, the identification of 24-hour rhythmic genes suggests potential compensation by the light-response pathway. However, most of rhythmic genes in the NvClk1-/- compared to WT polyps in LD conditions are different. This suggests the light-response pathway may not fully reproduce the normal circadian pattern observed in WT polyps and would require further investigation to understand their recruitment and function. Furthermore, the identification of a reduced number of rhythmic genes in DD NvClk1-/- polyps revealed the importance of NvClk in maintaining molecular rhythm in absence of light. However, the persistence of these rhythmic genes raises questions regarding their origin. They could be the result of the mutation itself, a compensatory mechanism or false positives resulting from our analysis. The study by Ray et al., which report controversial findings regarding RNA rhythmicity in the Bmal1-/- mouse cells (NvCycle orthologous in Nematostella) is an interesting case of rhythmicity analysis to consider34–38. Our attempt to identify rhythmic genes faced challenges, particularly in the choice of statistical parameters. Initially, using stringent criteria like BHQ for the JTK method or p.adj with RAIN resulted in a very limited number of genes for analysis (Supp. Table 3). Consequently, we opted for a balanced approach combining both algorithms with a p<0.01 threshold. We made a compromise to identify rhythmic genes with greater confidence, but we are aware that the methodological choices are critical for the results.

The clustering analysis revealed that rhythmic genes can be categorized by expression pattern similarity. This approach defined “temporal gene clusters” as group of seven/eight genes in average in the WT (Fig. 6b). Their discovery poses a fundamental question: How a group of genes are co-regulated in times and space (cell types) by the pacemaker? Their recruitment is disrupted in the DD NvClk1-/- polyps suggesting an essential function of NvClk in absence of light. The combination of published scAtlas39 and multiplexed FISH techniques40 will be essential to further investigate the biological regulation and function of these temporal gene clusters.

NvClk temporally organize pacemaker genes expression to drive rhythmic genes recruitment

Our study reveals that NvClk plays a key role in regulating the temporal transcription of pacemaker candidate genes (Fig.4a). Our analysis identified two clusters of pacemaker genes: One containing NvClk, and a second one containing a potential NvClk inhibitor (NvCipc)41,42. The alteration of clusters composition with a weaker anticorrelation in LD NvClk1-/- polyps might generate a desynchronization of the pacemaker factors availability. Indeed, regulation of rhythmic transcription involved a complex protein-protein-DNA timing interaction. Furthermore, we did not identify any circadian E-boxes enrichment in rhythmic genes between conditions, except for the candidate pacemaker genes. Altogether, this supports the function of NvClk in orchestrating the timing interaction of pacemaker factors to select downstream rhythmic genes, indicating a more complex regulatory landscape at play.

However, one significant unanswered question in our study is the reason for the arrhythmic transcription of putative pacemaker genes in DD. We hypothesize that using whole animals for sampling material might mask oscillating gene expression signals, especially if signals are present in a small number of cells or if tissues exhibit rhythmic gene expression in different phases. Furthermore, it is important to acknowledge a limitation in our interpretation, common in chronobiology: the use of RNA oscillation as a proxy for protein oscillation and function. The development of tools to study the pacemaker factors at the protein level in Nematostella will leverage this limitation in the field.

NvClk regulates processes involved in cell proliferation and the neural system in absence of light

Our study of NvClk suggests coordination of cellular processes, especially in the absence of light. The results of our rhythmic transcriptomic analysis (Fig. 2-3) raised questions regarding indirect effects, and the non-rhythmic function of NvClk. We performed a differential gene expression analysis on the total transcriptome for each light condition. Under LD conditions, NvClk1-/-polyps exhibited significant changes in gene expression but no GO term enrichment was found (Fig. 5a, Supp. Table 5) revealing multiple altered processes we cannot yet identified. In contrast, under DD conditions, NvClk1-/- polyps displayed more pronounced alterations, with more DEGs and enriched GO-terms for down-regulated genes related to mitosis, microtubule organization, and ciliary/flagellar motility, while the up-regulated genes showed enrichment in processes such as the modulation of other organism’s processes, axonal guidance, and sensory perception (Fig. 5b, Supp. Table 5). These results imply that NvClk has non-circadian functions that are dependent on light availability. This is particularly noteworthy considering the expression of core pacemaker genes, which are known to be arrhythmic during larvae stages, potentially involved in developmental processes22.

Taken as a whole, this study provides novel insights into circadian regulation in Nematostella vectensis and sheds light on the evolutionary origin of circadian time maintenance. Our findings indicate that CLOCK function is conserved from Cnidaria to mammals to maintain rhythmicity in absence of light cues. Furthermore, it revealed a light-response pathway able to compensate at both behavioral and molecular level using light cues. This circadian clock mutant opens new avenues for investigating cell-type-specific mechanisms of the circadian clock that drive the molecular and phenotypical oscillations of cnidarians. By further exploring the circadian clock mechanisms in cnidarians, we can gain deeper insights into the evolutionary origins of this critical aspect of biology, enhancing our understanding of how organisms have evolved to keep track of time and adapt to their environment.

Supporting information

eLife Assessment

This fundamental study for the first time defines genetically the role of the Clock gene in basal metazoa, using the cnidarian Nematostella vectensis. With convincing evidence, the study provides insight into the early evolution of circadian clocks. Clock in this species is necessary for daily rhythms under constant conditions, but not under a rhythmic light/dark cycle, suggesting that the major role of the circadian oscillator in this species could be a stabilizing function under non-rhythmic environmental conditions.

Reviewer #2 (Public Review):

In this revised manuscript Aguillon and collaborators convincingly demonstrating that CLK is required for free-running behavioral rhythms under constant conditions in the Cnidarian Nematostella. The results also convincingly show that CLK impacts rhythmic gene expression in this organism. This original work thus demonstrate that CLK was recruited very early during animal evolution in the circadian clock mechanism to optimize behavior and gene expression with the time-of-day. The manuscript could still benefit from some improvements so that it is more accessible for a wide readership.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #2 (Public Review):

In this revised manuscript Aguillon and collaborators convincingly demonstrating that CLK is required for free-running behavioral rhythms under constant conditions in the Cnidarian Nematostella. The results also convincingly show that CLK impacts rhythmic gene expression in this organism. This original work thus demonstrates that CLK was recruited very early during animal evolution in the circadian clock mechanism to optimize behavior and gene expression with the time-of-day. The manuscript could still benefit from some improvements so that it is more accessible for a wide readership.

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

Aguillon and collaborators have deeply revised, and in the progress significantly improved the presentation of their interesting results with the first Cnidarian circadian gene mutant. Results are now very convincingly demonstrating that CLK is required for free-running behavioral rhythms under constant conditions. The results also now more convincingly show that CLK impact rhythmic gene expression, although interpretation of the transcriptomics data is not straightforward. I think there is still improvements that are needed to make the manuscript more accessible. We authors need to keep in mind that a broad audience will read their report, not just chronobiologists. I have listed below several issues that I think should be addressed, and some editing suggestions.

General comment to Editor and Reviewers:

We are genuinely grateful to both reviewers and editors about all the feedback which helped us to make the best of our data, to question our analysis to the point we redefined our approach and end up with a great article we are proud of it. Only the name of authors is visible on the article, and considering how much the reviewing system help to improve the research it seems almost unfair. As such, we thank all of you and really appreciate the new eLife system. Bravo all.

Abstract:

(1) Line 40" It should read "transcript levels" instead of "transcription". There is no measurement of transcription rates in this manuscript, only mRNA levels.

Modified accordingly.

(2) Line 41: the authors mention "constant light". Does this refer to previous work? Their data in Figure 4 were in constant darkness, not in LL.

Modified accordingly.

(3) Line 46 and throughout the manuscript, the allelic nomenclature is not standard. 1-/- seems to indicate there are two different alleles. Since the allele might not be a null, I would suggest simply using 1/1, or perhaps delta/delta since the mutation results in a truncates CLK.

NvClk1-/- became NvClkΔ/Δ. Except in the .xls supplementary table were the mutant kept the NvClk-/- nomenclature. It is not possible to replace only part of a word with a different font, here generating delta sign would require to do it one by one.

(4) The last sentence of the abstract needs to be rephrased, as it suggests that CLK evolved to maintain circadian rhythms under constant conditions. Constant conditions very rarely exist on Earth, and thus cannot be an evolutionary driving force. Different explanations have been proposed on why a self-sustained clock is the evolutionary solution to timekeeping, but the purpose of the clock and of clock genes is not to maintain oscillations in constant conditions. Actually, this sentence conflicts with the title.

Modified to: the Clock gene has evolved in cnidarians to sustain 24-hour rhythmic physiology and behavior in absence of diel environmental conditions. From my actual understanding, you are right, the purpose of clock gene is not to maintain oscillation in constant conditions (this is simply the result of the experiment), but to synchronize the physiology to the day/night rhythm, and surely to sustain 24h oscillations in case the environment challenges the perception of the diel cues. The DD or LL is just an artificial experimental design to reveal the endogenous time-keeping pacemaker.

Results:

(1) Line 148 and elsewhere in the MS: I would not use the word "lower" or "higher" to qualify acrophases. I would suggest advanced/delayed or earlier/later.

Modified accordingly.

(2) Line 157-9: The introductory sentence does not clearly present the rationale for the 6/6 experiments.

We modified the paragraph accordingly: The presence of a 24-hour rhythm of NvClkΔ/Δ polyps under LD conditions could be attributed to either a direct light-response or the partial functioning of the circadian clock due to the nature of the mutation….

(3) At the end of the behavior section, or perhaps at the end of each paragraph in this section, it would be helpful to have a summary of the results and more clearly explain their interpretation. The authors need to guide the readers, particularly non-chronobiologist, so that they can understand what the really neat data that were obtained mean. For example, what does it mean that the acrophase is different between mutant and wild-type, why are Clk mutants rhythmic under LD12/12 or 6/6, etc.

We added a conclusion sentence to help non-specialist to understand each result.

(4) Line 172 and elsewhere" "true rhythmic genes" sounds odd to me. Either they are, or they are not rhythmic.

Modified to “rhythmic genes.”

(5) Paragraph starting with line 184: I do not follow what is important about the number of genes per time cluster. What does it tell us, beyond the simple fact that less genes are rhythmic in the Clk mutants?

We rewrote the result paragraph to make it clearer why we performed this clustering analysis. This clustering analysis became Extended Data Fig.2 with modification of the figures (see my comments in your review about Figure 3).

(6) Line 197: The authors need to explain what they saw with circadian clock genes and their expression in CLk mutants. In some case, amplitude increased in LD. This surprising observation deserves some explanations. "Complex regulatory effect" is too vague.

We replaced the vague “complex regulatory effect” by a more thorough description of the figure 3.a.

(7) Line 198-203: Again, help the reader understand the significance of these observations.

We rewrote the paragraph to help the reader to better understand the significance of these observations.

Discussion:

(1) Line 236-40. Careful with the use of -/-, which implies that an allele is a null. The first CLk mutants in mammals and flies, which the authors refer to. were actually dominant negatives.

I went over the citations we used for this paragraph and this first mutation in fly dClkar is null, no dominant negative. Flies are still rhythmic in the dark. Unless there is an older mutation? However, you right the first mutation identified in mouse was a dominant-negative with loss of rhythmicity, while the gene deletion did not show any effect on the behavior, suggesting compensation by a paralog. I removed two references which were not relevant to the discussion.

(2) Line 265-268 are not very clear. Do the authors mean that the lack of overlap for non-cricadian pacemaker genes is because of different experimental conditions? What would be those differences? It is reassuring that the Leach/Reitzel study and the present share pacemaker genes as rhythmic, but it is also surprising that there is almost no overlap beyond these genes. How robust are those other rhythms compared to circadian clock genes?

We revised this paragraph and raised major points regarding the raising condition of our polyps between labs and their potential genetic differences which could explain these differences.

(3) Line 270. I am not sure "compensation" is the right word, since there is no overlap between the rhythmic genes in mutants under LD and wild-type under either LD or DD. Also, saying on line 273 that the transcriptional pattern is not fully reproduced is a rather striking understatement, given the absence of rhythm gene overlap

We rewrote the paragraph accordingly. We replaced by “alternative way to drive rhythmicity under LD condition”.

(4) Line 279. The authors mention the possibility of false positives. Based on the FDR, is there more rhythmic genes than by chance?

The possibility of false-positive is a risk to consider when you do not perform multiple-testing. We added within the results paragraph the number of rhythmic genes identified with BH.Q or p.adj. which both are the multiple testing for each algorithm (RAIN and JTK) we used.

(5) Line 279-82. The references to the Ray study is rather obscure. What is the point the authors are trying to make here?

Eventually, we removed the reference from this article and modify the paragraph of the discussion. Indeed, the discussion around the Ray study did not gave an interesting direction to discuss our results and analysis approach.

(6) Line 284: define BHQ and p.adj

Defined and referenced.

(7) The way Lines 283-288 are worded do not provide a good rationale for how transcriptional rhythms were analyzed. The idea to combine two different approaches (JTK and RAIN) to be selective with rhythmicity was great. The authors need however to justify these choices in a more convincing manner. The goal is to detect rhythmic genes in a reliable manner, irrespective of the number of rhythmic genes observed Also, explaining the choice of methodology belongs to the result section.

We explained our choice of methodology and moved it to the result section as suggested.

(8) Line 292-3. There are known mechanisms that explain how transcriptional time clusters are generated. In particular, the use of interlocked feedback loop with antiphase peaks of transcriptions is well documented. Actually, it seems to me the clustering shown in Fig 4 might hint at such a mechanism.

Indeed you are right the clustering shown in Fig 3 (former Fig 4) revealed such mechanism.

Figures:

Figure 2: Define relative amplitude

We added the definition of the relative amplitude within the results. If this is what you asked for?

Figure 3: Some of the cycles look odd (first row of graphs in panel C). Why would the first and last data point be so different in three of these graphs?

We decided to modify this figure as we realized it was not informative and not objective enough, as we selected among multiple patterns few “representatives”. In the new figure we combined the cluster analysis to the behavior. Thus, readers can now pick a cluster according to a specific behavior activity level (or ZT/CT) and reach in supp. Table 4 the “genes of potential interest”. However generally speaking this figure does not explain more the consequences of the mutation, so we moved it into the Extended data Fig.2

Figure4: define the color coding in the correlation panels (blue to red)

These values from -1 to 1 are the Pearson correlation values. Now indicated on the figure with the color coding legend.

Version 3

Abstract

Abstract

The circadian clock enables anticipation of the day/night cycle in animals ranging from cnidarians to mammals. Circadian rhythms are generated through a transcription-translation feedback loop (TTFL or pacemaker) with CLOCK as a conserved positive factor in animals. However, CLOCK’s functional evolutionary origin and mechanism of action in basal animals are unknown. In the cnidarian Nematostella vectensis, pacemaker gene transcript levels, including NvClk (the Clock ortholog), appear arrhythmic under constant darkness, questioning the role of NvCLK. Utilizing CRISPR/Cas9, we generated a NvClk allele mutant (NvClkΔ), revealing circadian behavior loss in constant light conditions (LL and DD) while a 24-hour rhythm was maintained under light-dark condition (LD). Transcriptomics showed distinct rhythmic genes in wild-type (WT) genes in LD compared to DD. The LD NvClkΔ/Δ showed comparable numbers of rhythmic genes, but were greatly reduced in DD. Furthermore, the LD NvClkΔ/Δ showed alterations in temporal pacemaker gene expression, affecting their potential interactions. Additionally, differential expression of non-rhythmic genes associated with cell division and neuronal differentiation was observed. These findings revealed that a light-responsive pathway can partially compensate for circadian clock disruption and that the Clock gene has evolved in cnidarians to synchronize rhythmic physiology and behavior to the diel rhythm of the earth’s biosphere.

Introduction

Throughout the history of life on Earth, organisms have had to adapt to a constantly changing environment, including the ∼24-hour daily rhythm of light/dark, driving the development of endogenous biological clocks. The circadian clock, entrained by external stimuli such as light, enables the organism to anticipate the onset of the light and dark phases and synchronize its physiology and behavior in harmony with the environment. This, in turn, enhances the organism’s fitness and survival1–3. From single-celled organisms to metazoans, circadian clocks have evolved multiple times, highlighting their importance to living organisms1,2. Despite the fundamental role circadian clocks play in regulating the rhythmicity of living organisms, their evolutionary origin and intricate molecular mechanisms remain ambiguous in early diverging animal lineages, such as cnidaria.

Rhythmic phenomena, including calcification, reproduction, and diel behavior patterns, have been examined in cnidarian species3–8. While environmental stimuli, such as light, directly triggered some of these patterns, others persist in the absence of external cues, suggesting the presence of an internally generated and self-sustaining circadian clock6,9,10. At the molecular level, cnidarians possess homologs of putative core pacemaker genes found in bilaterians11–13. Several studies have shown that most of these genes display diel expression patterns under light/dark cycles. However, unlike most animals, their oscillation generally ceases without light cues14–16. Thus, how the core pacemaker genes orchestrate rhythmic gene expression and circadian behaviors in cnidarians remains unclear.

One of the most studied cnidarian species in the field of chronobiology is the estuarine sea anemone, Nematostella vectensis. Few studies have shown that in diel lighting, the locomotor behavior of Nematostella has a ∼24-h rhythm that is maintained under constant conditions, suggesting it is regulated by an endogenous circadian clock6,17–19. By this, the Nematostella genome codes for conserved core pacemaker genes such as NvClk, NvCycle, and the cryptochromes NvCry1a and NvCry1b12,13. The proposed circadian clock model in Nematostella is composed of the positive transcription factors (bHLH-PAS family), NvCLK and NvCYCLE, that heterodimerize and upregulate light-dependent cryptochrome genes in the feedback loop, and NvPAR-bZIPs in the feed-forward loop, which repress the transcription of the positive elements11,12,19. More recently, the NvCLK-interacting pacemaker, NvCIPC, was predicted to act as an additional repressor of the NvCLK:NvCYCLE dimer19. However, in contrast to the free-running oscillation demonstrated for Nematostella behavior6,17,19, transcriptional expression profiles of most candidate genes implicated in the pacemaker do not retain their oscillation period without light14,15,19.

We employed the CRISPR/Cas9-mediated genome editing system to establish a NvClk mutant (NvClkΔ/Δ) Nematostella. By combining behavioral monitoring and transcriptomic analysis, we aimed to elucidate the role of NvClk in regulating rhythmic locomotor activity and gene expression under varying light conditions. Our study revealed a robust light response pathway capable of compensation and a conserved function of CLOCK as a timekeeper without a light cue.

Results

Phylogenetic analysis and spatial expression pattern of NvClk

Phylogenetic analysis of NvCLK protein sequences positioned NvClk within the cnidarian branch (Fig. 1a). It contains a basic helix–loop–helix (bHLH) DNA binding domain and two Per-Arnt-Sim (PAS) domains, similar to the protein structure found in other animals. PAS domains are crucial structural motifs in protein-protein interactions that drive the self-sustaining molecular mechanism underlying the circadian clock20,21.

NvClkΔ/Δ cannot maintain circadian behavior in non-diel light conditions.

(a) Phylogenetic tree showing the evolutionary relationship of CLK orthologs across different animal species. (b) In situ hybridization of NvClk in the WT juvenile with scale bars representing 0.1mm. (c) Schematic representation of the NvClk gene in grey, with the open reading frame (ORF) in dark grey and the conserved protein domains bHLH (yellow) and PAS1 and PAS2 (dark red). The CRISPR-generated NvClkΔ allele has a +20nt insertion after the PAS1 domain, represented by a black arrowhead. NvCLK dimerizes via its three functional domains with NvCYCLE binding the CACGTG ebox to drive rhythmic transcription. (d-g-j-m). Normalized Movement (a.u), hourly binned over 72h, under different light conditions: 12h light:12h dark, continuous dark (Dark - Dark), continuous light (Light - Light), and 6h light:6h dark. The black line represents the WT, and the red line represents the NvClkΔ/Δ mutant. (e-h-k-n) Lomb-Scargle Periodograms for each corresponding light condition. The significant period value (p<0.01) is indicated for each genotype in the top left corner of each graph. (f-i-l-o) Phase detection (Cosinor) and genotype comparison of 24h rhythmic individuals. See the number rhythmic/total on the x-axis indicating the number of 24h-rhythmic animals over the total population for each genotype.

In situ hybridization chain reaction (HCRv.3) was performed to localize NvClk expression at the polyp stage. Polyps were sampled at ZT10, i.e., peak expression of NvClk6,12,22. NvClk mRNA expression was observed throughout the animal tissue, and enriched expression was visible in the tentacle endodermis and mesenteries. In contrast, no signal was observed in the negative control (Fig.1b, Extended Data Fig.1a). This expression pattern resembled the expression observed at the larvae stage22. To date, functional manipulation of the NvClk gene has not been performed in basal animal lineages including, cnidarians, and its function is unknown in cnidarians20,21 (Fig.1a).

Generation of NvClkΔ/Δ Nematostella

To investigate the function of NvClk in Nematostella, we employed the CRISPR-Cas9 system to generate mutants. Based on existing knowledge from mouse and Drosophila models, we hypothesized that NvCLK:NvCYCLE dimer binds to the DNA motif CACGTG within the promoter of rhythmic target genes (Fig. 1c). Guide RNA (gRNA) was synthesized to target a region between the two PAS domains of the NvClk coding sequence (CDS). This gRNA and the Cas9 endonuclease were microinjected into zygotes (Methods). Subsequently, F0 animals were outcrossed with wild type (WT), and the F1 progeny were raised to adulthood. Genotyping of F1 polyps identified ten different mutated alleles, with six displaying a frame-shift mutation, including one with a 20 bp insertion (NvClkΔ), resulting in a premature stop codon (Extended Data Fig. 1b). The predicted 203 amino acid truncated protein lacked 459 amino acids, including one co-factor dimerization PAS domain (Fig. 1c, Extended Data Fig. 1b). To obtain homozygous NvClkΔ/Δ polyps, we crossed heterozygous NvClkΔ F1 animals. Genotyping of F2 polyps confirmed the expected 25% frequency of NvClkΔ/Δ mutants. Subsequently, we intercrossed NvClkΔ/Δ animals to obtain F3 NvClkΔ/Δ polyps for use in subsequent experiments aimed at assessing the role of NvClk in regulating behavioral and genetic rhythms.

NvClk is necessary to maintain circadian behavior in constant conditions

To assess the impact of the NvClkΔ mutation on circadian rhythm, we monitored the locomotor behavior of WT and NvClkΔ/Δ polyps under different light conditions (Supp. Table 1). Both the WT and NvClkΔ/Δ populations exhibited a 24-hour periodicity in 12:12 hour Light-Dark (LD) cycles (Fig. 1d, e), with 15 out of 21 WT animals displaying 24-hour rhythmicity compared to only 9 out of 20 NvClkΔ/Δ animals (Fig. 1f, Table 1, Supp. Table 2). The average acrophase for WT polyps (13.3 hours) was significantly earlier than for NvClkΔ/Δ polyps (17.3 hours) (Fig. 1f, Table 1). While we could detect a 24-hour rhythm for both genotypes, the delayed acrophase and reduced number of significant rhythmic polyps in the NvClkΔ/Δ suggest an alteration of the underlying rhythmicity mechanism.

We then investigated locomotor behavior under continuous conditions, namely continuous dark (DD) or continuous light (LL). WT polyps exhibited a 22-hour rhythmic behavior under both constant light conditions, with 17 out of 25 WT polyps displaying a 24-hour rhythm under DD and 7 out of 25 under LL (Fig. 1g-l). In contrast, a few NvClkΔ/Δ polyps displayed rhythmic behavior under constant conditions (1 out of 24 in DD and 1 out of 26 in LL) (Table 1). Additionally, we observed an intermediate phenotype in the locomotor behavior of heterozygous polyps for the NvClkΔ allele in DD (Extended Data Fig. 1c-f). These results revealed that NvClkΔ/Δ polyps could not maintain circadian rhythmicity without diel light cues.

A 24-hour rhythm of NvClkΔ/Δ polyps under LD conditions could be attributed to either direct light response or the partial functioning of the circadian clock due to the nature of the mutation. To distinguish between these two possibilities, we monitored locomotor activity under a 6-hour light: 6-hour dark (LD 6:6) cycle after a regular diel 72-hour entrainment under 12:12 LD. While WT polyps maintained a marginally significant periodicity of 22 hours, NvClkΔ/Δ polyps displayed a 12-hour rhythm at the population level (Fig. 1m-o). Specifically, we identified a clear difference of 12-hour rhythmic individual polyps between WT and NvClkΔ/Δ groups (1 out of 25 WT polyps vs. 13 out of 26 NvClkΔ/Δ polyps) (Table 1). Notably, entrainment with LD 6:6 did not lead to a 12-hour rhythm in DD for both WT and NvClkΔ/Δ polyps (Extended Data Fig. 1g-i). These results support the hypothesis that the 24-hour rhythm observed in the NvClkΔ/Δ polyps in LD condition is due to the light-response pathway and not from an endogenous oscillator.

NvClk regulates rhythmic gene expression differentially in response to light conditions

We conducted transcriptional profiling to investigate the underlying molecular correlates of the behavioral phenotype found in NvClkΔ/Δ polyps. WT and NvClkΔ/Δ polyps were sampled seven times every 4 hours over 24 hours under LD and DD conditions (Fig. 2a). To identify rhythmic genes, we employed stringent statistical parameters, including Benjamini-Hochberg (BH.Q) for the JTK method23 and adjusted p-value (p.adj) for the RAIN method24. This resulted in the identification of a minimal number of rhythmic genes. We detected only six rhythmic genes under LD conditions in WT polyps using the JTK method and 40 rhythmic genes using the RAIN method (Supp. Table 3). In DD condition, in the WT polyps, only two rhythmic genes were identified using the RAIN method (Supp. Table 3). Despite the risk of false positives, we opted not to use multiple testing but instead proposed to combine the JTK and RAIN algorithms to identify rhythmic genes, ensuring a robust approach to data analysis (p<0.01). We identified 119 rhythmic genes rhythmic under LD and 107 rhythmic genes under DD in WT polyps (Fig. 2b, Supp. Table 3). In NvClkΔ/Δ polyps, we detected 147 rhythmic genes under LD and only 37 under DD (Fig. 2b, Supp. Table 3).

NvClkΔ/Δ shows rhythmic gene reduction in constant darkness with altered rhythmic features.

(a) Overview of the experimental design used to generate RNA-seq data. Polyps were entrained for 72 hours before sampling at 4-hour intervals over a 24-hour period (dark arrows) in both LD and DD cycles. (b) Venn diagram comparing the total number of 24h rhythmic genes identified in WT and NvClkΔ/Δ in LD and DD cycles with a p <0.01 with RAIN and JTK. (c) Average acrophase comparison between rhythmic genes in LD and DD in WT polyps. Mann-Whitney test, p<0.001. (d) Average acrophase comparison between rhythmic genes in LD and DD in NvClkΔ/Δ polyps. Mann-Whitney test, p>0.05. (e) Average relative amplitude comparison between rhythmic genes in LD and DD in WT polyps. Mann-Whitney test, p<0.05. (d) Average relative amplitude comparison between rhythmic genes in LD and DD in NvClkΔ/Δ polyps. Mann-Whitney test, p>0.05. (c-f) sample size (n) indicated below each boxplot.

The rhythmic genes in WT polyps displayed a delayed acrophase under DD compared to LD (17.20h vs. 12.93h, Fig. 2c). However, no differences were detected between LD and DD rhythmic genes in NvClkΔ/Δ polyps (Fig. 2d). Similarly, the relative amplitude (the gene amplitude divided by its baseline, rAMP) of DD rhythmic genes was higher in WT polyps compared to LD (0.61 vs. 0.43, Fig. 2e), but no rAMP difference was observed between LD and DD rhythmic genes in NvClkΔ/Δ polyps (Fig. 2f).

Are rhythmic genes organized into “transcriptional time clusters”? Does the NvClkΔ mutation modify cluster recruitments, causing the loss of rhythmic behavior under DD conditions? We performed a clustering analysis on the rhythmic genes using the DPGP model (Dirichlet process Gaussian process mixture model). The number of genes per cluster between LD and DD conditions in WT polyps did not differ significantly (7.3 vs. 7.6, Extended Data Fig. 2a, Supp. Table 4). Interestingly, when clusters are organized by their acrophase, we observed clusters with higher numbers of genes peaking at subjective night in WT under DD conditions (Extended Data Fig. 2b, Supp. Table 4). In NvClkΔ/Δ polyps, the number of genes per cluster was significantly reduced in DD compared to the LD condition (4.1 vs. 8.6, Extended Data Fig. 2c,d). We did not identify GO-term enrichment in any cluster. However, the overlap between clusters and behavior opens new directions for further functional analysis (Extended Data Fig. 3b,d and Supp. Table 4). Overall, the reduced number of rhythmic genes in NvClkΔ/Δ polyps under the DD condition and the reduced number of genes per cluster confirm the necessity of NvClk to recruit rhythmic genes in the DD condition and to organize them in transcriptional time clusters.

NvClk regulates the temporal expression pattern of pacemaker genes

In line with previous findings in Nematostella12,14, candidate pacemaker genes showed arrhythmic expression under DD conditions (Fig. 3a, Supp. Table 3). However, the altered expression patterns observed in NvClkΔ/Δ compared to WT polyps in LD condition showed increased transcripts for some genes (i.e., NvClk and NvPar-bzipd). In contrast, others (NvCipc and NvPar-bzipc) exhibited a reduction in transcript numbers (Fig. 3a, Supp. Table 3). If we hypothesize that the first two genes (NvClk and NvPar-bzipd) act as positive factors and the latter two (NvCipc and NvPar-bzipc) potentially serve as negative regulator of the former, the lack of functionality of the NvClkΔ allele would explain the observed difference in transcript levels between NvClkΔ/Δ and WT polyps.

NvClkΔ/Δ alters temporal pacemaker gene expression.

(a) Four pacemaker genes are plotted, showing the read counts over 24h in LD and DD in WT (black) and NvClkΔ/Δ (red). The continuous line represents significant rhythmicity (RAIN&JTK p<0.01), while the dashed line indicates no rhythmicity. (b) correlation matrix of candidate pacemaker genes expression in LD for WT on the left and NvClkΔ/Δ on the right. (c) schematic representation of the promoter sequences analyses 5kb upstream of the putative ATG. Black boxes represent canonical E-boxes, while circadian E-boxes are green. Below is the logo motif we used to identify canonical and circadian Ebox. (d) Circadian / Canonical ratio (in %) per condition. Kruskal-Wallis, multiple comparison, a vs b : p<0.05.

To systematically assess the mutation’s impact on all the potential pacemaker genes, we utilized a correlation matrix based on their temporal transcript number levels, offering a comprehensive overview of their temporal organization. In WT polyps under LD conditions, the clustering categorized genes into two groups: one exhibiting a daytime peaking, containing NvClk, and another peaking at night comprising NvPar-bzipc and NvCipc. Notably, in LD NvClkΔ/Δ polyps, this second cluster contained two additional genes and displayed a weakened anticorrelation with the NvClk cluster (Fig. 3b). These observations suggest that the pacemaker oscillation, generated by the interplay of positive and negative feedback loops, relies on the precise temporal organization of these potential pacemaker factors into distinct clusters. The disruption of this organization by the NvClkΔ allele underscores the central role of NvClk in pacemaker function.

To go further into the regulatory mechanisms downstream of the pacemaker, we examined the presence of circadian E-box motifs (CACGTG) within 5kb upstream of the predicted ATG of rhythmic genes. We calculated circadian/canonical E-box enrichment to account for the total variation in the number of canonical E-boxes (Fig. 3c). Notably, only the candidate pacemaker genes exhibited a significant enrichment in circadian E-boxes in their promoters (15.9%) compared to the WT (5.6%), NvClkΔ/Δ (4.8%) rhythmic genes, and non-rhythmic genes (6.8%) (Fig. 3d).

NvClk coordinates cell division and neuronal pathways in constant darkness

In addition to the transcriptomic rhythmic analysis, we aimed to identify processes regulated by NvClk that may not necessarily exhibit rhythmicity. We conducted a differential gene expression analysis on the total transcriptome between genotypes under each light condition to achieve this. Under LD conditions, NvClkΔ/Δ polyps exhibited 457 down-regulated genes and 646 up-regulated genes, with no significant enrichment in GO terms observed (Fig. 4a, Supp. Table 4 and 5). However, in DD conditions, NvClkΔ/Δ displayed 2450 down-regulated genes and 1770 up-regulated genes (Fig. 4b, Supp. Table 4). Notably, we identified enrichment in down-regulated genes in processes related to mitosis, microtubules, and ciliary/flagellar motility. Conversely, the up-regulated genes showed significant enrichment in processes such as the modulation of another organism’s processes, axonal guidance, and sensory perception (Fig. 4b, Supp. Table 5).

NvClkΔ/Δ disrupts cell-cycle and neuronal pathways in constant darkness.

(a) Volcano plot showing the differential expression of genes (DEG) between WT and NvClkΔ/Δ in LD (Left) and DD (right). Dashed line indicates the threshold used to detect DEG (p.adj<0.01). Red dots indicate down regulated genes and black dots up-regulated genes in NvClkΔ/Δ compare to WT polyps (b) Gene Ontology (GO) terms with with significant fold-enrichment (Bonferroni corrected p-value or p.adjusted <0.01) for the DEG analysis in DD. Down regulated genes in Red while Up regulated genes in Black.

Discussion

Conserved behavioral CLOCK function through animal evolution

Our study provides valuable insights into the evolution of circadian clocks by characterizing the effects of the first Clock mutation in a cnidarian, the sea anemone Nematostella vectensis. Our behavioral assays showed that NvClk is essential for maintaining rhythmic locomotor activity without an entraining light cue. Although the rhythmicity of the NvClk+/Δ heterozygote polyps was affected in DD, our results could not discriminate a dominant-negative from a total loss of function to identify the nature of this mutation (Extended Data Fig. 1g-i). Studies in various model organisms further support the importance of CLOCK in regulating circadian locomotion. For instance, DmClkar/ar flies exhibit a loss of circadian locomotion in constant darkness25, while the DnClk1adg3/dg3 zebrafish mutant displayed a shortened period under the same conditions26. The dominant-negative mutant MmClock Δ5–6/Δ5–6 mice showed a loss of circadian locomotion in constant darkness, however the complete deletion of the MmClock gene did not affect the circadian behavior rhythm in constant darkness suggesting compensation by a paralog27–29. Overall, these findings support a conserved role of CLOCK in preserving circadian behavioral rhythms in the absence of light cues across the distant Nematostella, flies, zebrafish, and mice.

Moreover, the conservation of a 24-hour locomotion rhythm in LD of the NvClkΔ/Δ polyps with a delayed acrophase revealed a light-response pathway independent of the circadian circuit, consistent with observations in other animal models25,27,30 (Fig. 1f). NvClkΔ/Δ polyps exposed to a 12:12h LD cycle exhibited a 24-hour period. In contrast, those exposed to a 6:6h LD cycle displayed a 12-hour period. Notably, nearly no WT polyps exhibited a 12-hour rhythm under this condition, suggesting that the circadian clock overrides the light-response pathway (Fig. 5a). While some of the circadian factors can directly sense the light, such as CRY proteins31, 29 putative NvOpsin have been identified in the genome which could be involved in the light-response pathway32. Behavioral tracking of NvClkΔ/Δ polyps exposed to different wavelengths could help to identify candidates for further functional studies of the light-response pathway.

Summary of NvClk function in the regulation of Nematostella circadian rhythmicity.

Transcriptional rhythmicity plasticity downstream NvClk

At the transcriptomic level, previous studies in Nematostella have shown large changes in the transcriptional profile of many genes after a single day of constant darkness, including the candidate pacemaker genes that were found arrhythmic despite sustaining circadian locomotion12,14,22. Consistent with previous transcriptomic analysis in cnidarian7,14,16,33, most of the rhythmic genes identified in LD differed from those identified in DD in the WT polyps. Notably, they displayed higher mean acrophase and larger mean amplitude in DD, suggesting a differential regulation in response to light conditions, which had not been investigated in previous cnidarian studies. Furthermore, the overlap observed between our LD rhythmic genes and those identified by Leach et al.14 underscores the robustness of pacemaker rhythmic transcription in LD conditions (Extended data Fig. 3). However, the lack of overlap for rhythmic genes raises intriguing questions. Differences in experimental conditions, including genetic backgrounds, light system (Neon vs. LED), salinity (12ppt vs. 15ppt), and temperature (17°C vs. 25°C), may contribute to these discrepancies. Further investigations are necessary to determine if the lack of overlap of rhythmic genes downstream of the potential pacemaker genes results from an organism’s adaptation to its environment and, therefore, reflects the plasticity of the pacemaker in regulating its downstream rhythmic genes.

Our study identified 24-hour rhythmic behavior in NvClkΔ/Δ polyps under LD conditions, suggesting an alternative mechanism for generating molecular rhythmicity via the light-sensing pathway. However, it is crucial to note minimal overlap between the rhythmic genes identified in NvClkΔ/Δ and WT polyps under LD conditions. This discrepancy indicates that the light-response pathway may not fully replicate the normal pacemaker functions observed in WT polyps, highlighting the need for further investigation into the recruitment and function of these genes. Additionally, the reduced number of rhythmic genes identified in NvClkΔ/Δ polyps under the DD condition underscores the crucial role of NvClk in maintaining molecular rhythm without light cues.

The clustering analysis revealed that rhythmic genes can be categorized into “transcriptional time clusters” (aka synexpression clusters)16,34 by group of seven/eight genes in average in the WT (Fig. 5b). Their existence raises a fundamental question that has yet to be answered: How is a group of genes co-regulated in time and space (cell types) by the pacemaker? Their recruitment is disrupted in the DD NvClkΔ/Δ polyps suggesting an essential function of NvClk in absence of light. The combination of published scAtlas35 and multiplexed FISH techniques36 will be essential to further investigate the biological regulation and function of these transcriptional time clusters.

NvClk temporally organizes pacemaker gene expression to drive rhythmic gene recruitment

Our study reveals that NvClk plays a crucial role in regulating the temporal transcription of pacemaker candidate genes (Fig.3a). Our analysis identified two clusters of pacemaker genes: One containing NvClk and a second one containing a potential NvClk inhibitor (NvCipc)37,38. These two clusters suggest the organization of the potential pacemaker genes transcription into interlock feedback loops with antiphase peaks, probably at the origin of the pacemaker oscillator function21,39,40. The alteration of cluster composition with a weaker anticorrelation in LD NvClkΔ/Δ polyps might generate a desynchronization of the pacemaker factors’ availability. Indeed, regulation of rhythmic transcription involved a complex protein-protein-DNA timing interaction. Furthermore, we did not identify any circadian E-boxes enrichment in rhythmic genes between conditions, except for the candidate pacemaker genes. Altogether, this supports the function of NvClk in orchestrating the timing interaction of pacemaker factors to select downstream rhythmic genes, indicating a more complex regulatory landscape at play.

However, one significant unanswered question in our study is the reason for the arrhythmic transcription of putative pacemaker genes in DD. Using whole animals for sampling material might mask oscillating gene expression signals, especially if signals are present in a small number of cells or if tissues exhibit rhythmic gene expression in different phases. Furthermore, we must acknowledge a limitation in our interpretation, which is common in chronobiology: using RNA oscillation as a proxy for protein oscillation and function. The development of tools to study the pacemaker factors at the protein level in Nematostella will leverage this limitation in the field.

NvClk regulates processes involved in cell proliferation and the neural system in the absence of light

Our study of NvClk suggests coordination of cellular processes, especially in the absence of light. Our rhythmic transcriptomic analysis results (Fig. 2 and Extended data Fig. 2) raised questions regarding indirect effects and the non-rhythmic function of NvClk. We performed a differential gene expression analysis on the total transcriptome for each light condition. Under LD conditions, while NvClkΔ/Δ polyps exhibited significant changes in gene expression, we could not identify any GO term enrichment (Fig. 4a, Supp. Table 5), revealing multiple altered processes we cannot yet identify.

In contrast, under DD conditions, NvClkΔ/Δ polyps displayed more pronounced alterations, with more DEGs and enriched GO-terms for down-regulated genes related to mitosis, microtubule organization, and ciliary/flagellar motility, while the up-regulated genes showed enrichment in processes such as the modulation of other organism’s processes, axonal guidance, and sensory perception (Fig. 4b, Supp. Table 5). These results imply that NvClk has non-circadian functions dependent on light availability. This is particularly noteworthy considering the expression of core pacemaker genes, known to be arrhythmic during larvae stages, potentially involved in developmental processes22.

This study provides novel insights into circadian regulation in Nematostella vectensis and sheds light on the evolutionary origin of circadian time maintenance. Our findings indicate that CLOCK function is conserved from cnidaria to mammals to maintain rhythmicity without diel light cues. Furthermore, it revealed a light-response pathway able to compensate at both behavioral and molecular levels using light cues. This circadian clock mutant opens new avenues for investigating cell-type-specific mechanisms of the circadian clock that drive the molecular and phenotypical oscillations of cnidarians. By further exploring the circadian clock mechanisms in cnidarians, we can gain deeper insights into the evolutionary origins of this critical aspect of biology, enhancing our understanding of how organisms have evolved to keep track of time and adapt to their environment.

Supporting information